---
title: "cdc_logistic_regression"
author: "Zach Skiba"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Section 0 — Setup & Package Loading

## Overview

We load all required packages and set a global seed for reproducibility.
Key packages: `caret` for cross-validation and training, `glmnet` for
regularized logistic regression, `nnet` for multinomial logistic
regression, `themis` for SMOTE inside CV folds, `pROC` for ROC-AUC, and
`ggplot2` / `kableExtra` for visualization and reporting.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r load-packages}
required_packages <- c(
  "caret", "glmnet", "nnet", "themis", "pROC",
  "ggplot2", "kableExtra", "dplyr", "recipes", "tidyr",
  "MLmetrics"
)

installed  <- rownames(installed.packages())
to_install <- setdiff(required_packages, installed)
if (length(to_install) > 0) install.packages(to_install, dependencies = TRUE)

library(caret)
library(glmnet)
library(nnet)
library(themis)
library(pROC)
library(ggplot2)
library(kableExtra)
library(dplyr)
library(recipes)
library(tidyr)
library(MLmetrics)

set.seed(571)
```

# Section 1 — Data Loading & Validation

## Overview

We load all four pre-split CSVs directly — no re-splitting is done here.
The data was stratified 80/20 by a teammate. We verify dimensions,
column alignment, missing values, class distributions, and convert
target variables to factors, which `caret` requires for classification.

```{r load-data}
binary_train     <- read.csv("../data/processed/binary_train.csv")
binary_test      <- read.csv("../data/processed/binary_test.csv")
multiclass_train <- read.csv("../data/processed/multiclass_train.csv")
multiclass_test  <- read.csv("../data/processed/multiclass_test.csv")
```

```{r validate}
# Dimensions
cat("binary_train:    ", nrow(binary_train),     "x", ncol(binary_train),     "\n")
cat("binary_test:     ", nrow(binary_test),      "x", ncol(binary_test),      "\n")
cat("multiclass_train:", nrow(multiclass_train), "x", ncol(multiclass_train), "\n")
cat("multiclass_test: ", nrow(multiclass_test),  "x", ncol(multiclass_test),  "\n")

# Column alignment
cat("\nColumns match (binary):    ",
    identical(colnames(binary_train), colnames(binary_test)), "\n")
cat("Columns match (multiclass):",
    identical(colnames(multiclass_train), colnames(multiclass_test)), "\n")

# Missing values
cat("\nNA counts:\n")
cat("binary_train:",     sum(is.na(binary_train)),     "\n")
cat("binary_test:",      sum(is.na(binary_test)),      "\n")
cat("multiclass_train:", sum(is.na(multiclass_train)), "\n")
cat("multiclass_test:",  sum(is.na(multiclass_test)),  "\n")
```

```{r class-distribution}
# Binary class distribution
cat("Binary target distribution (train):\n")
print(round(prop.table(table(binary_train$Diabetes_binary)) * 100, 2))

# Multiclass class distribution
cat("\nMulticlass target distribution (train):\n")
print(round(prop.table(table(multiclass_train$Diabetes_012)) * 100, 2))
```

```{r convert-factors}
# caret requires factor targets; X prefix avoids issues with numeric level names
binary_train$Diabetes_binary  <- factor(binary_train$Diabetes_binary,
                                        levels = c(0,1), labels = c("X0","X1"))
binary_test$Diabetes_binary   <- factor(binary_test$Diabetes_binary,
                                        levels = c(0,1), labels = c("X0","X1"))

multiclass_train$Diabetes_012 <- factor(multiclass_train$Diabetes_012,
                                        levels = c(0,1,2), labels = c("X0","X1","X2"))
multiclass_test$Diabetes_012  <- factor(multiclass_test$Diabetes_012,
                                        levels = c(0,1,2), labels = c("X0","X1","X2"))

cat("Binary levels:    ", levels(binary_train$Diabetes_binary), "\n")
cat("Multiclass levels:", levels(multiclass_train$Diabetes_012), "\n")
```

# Section 2 — Preprocessing

## Overview

We apply z-score normalization to three continuous features: `BMI`,
`MentHlth`, and `PhysHlth`. Binary and ordinal features are left
unchanged.

The scaler is fit **only on training data**, then applied to both train
and test sets. This is not "touching" the test set in a leakage sense —
we never fit or learn anything from test data. We simply transform it
using parameters derived exclusively from training, so that the model
receives inputs on a consistent scale at evaluation time.

```{r preprocess}
scale_cols <- c("BMI", "MentHlth", "PhysHlth")

# Fit on binary training data only
train_means <- colMeans(binary_train[, scale_cols])
train_sds   <- apply(binary_train[, scale_cols], 2, sd)

# Apply to both binary sets using training parameters
binary_train[, scale_cols] <- scale(binary_train[, scale_cols],
                                    center = train_means,
                                    scale  = train_sds)
binary_test[, scale_cols]  <- scale(binary_test[, scale_cols],
                                    center = train_means,
                                    scale  = train_sds)

# Fit on multiclass training data only
mc_means <- colMeans(multiclass_train[, scale_cols])
mc_sds   <- apply(multiclass_train[, scale_cols], 2, sd)

# Apply to both multiclass sets using training parameters
multiclass_train[, scale_cols] <- scale(multiclass_train[, scale_cols],
                                        center = mc_means,
                                        scale  = mc_sds)
multiclass_test[, scale_cols]  <- scale(multiclass_test[, scale_cols],
                                        center = mc_means,
                                        scale  = mc_sds)
```

```{r validate-scaling}
cat("Binary train — mean and sd after scaling (expect ~0 and ~1):\n")
print(round(sapply(binary_train[, scale_cols], function(x)
  c(mean = mean(x), sd = sd(x))), 4))

cat("\nBinary test — mean and sd after scaling (will differ slightly, expected):\n")
print(round(sapply(binary_test[, scale_cols], function(x)
  c(mean = mean(x), sd = sd(x))), 4))
```

# Section 3 — Binary Logistic Regression

## Section 3a — Baseline glm() (No Imbalance Handling)

## Overview

We first fit a plain logistic regression with no imbalance handling and
evaluate it on the **training set only**. This is intentionally wrong —
evaluating on the data the model was trained on produces overly
optimistic accuracy and is not a valid performance estimate. We do this
purely as a reference point to show the limitation of naive accuracy as
a metric when classes are imbalanced. With \~86% of the training data
being class 0, a model that predicts 0 for everything would achieve
\~86% accuracy — this chunk exposes that problem.

```{r baseline-glm}
set.seed(571)

# Fit baseline logistic regression on full training set
baseline_glm <- glm(Diabetes_binary ~ .,
                    data   = binary_train,
                    family = binomial)

# Predict on training data — intentionally wrong, reference point only
train_probs <- predict(baseline_glm, binary_train, type = "response")
train_preds <- factor(ifelse(train_probs > 0.5, "X1", "X0"), levels = c("X0", "X1"))

# Full confusion matrix
cm <- confusionMatrix(train_preds, binary_train$Diabetes_binary, positive = "X1")
print(cm$table)

cat("\nAccuracy:    ", round(cm$overall["Accuracy"], 4), "\n")
cat("Sensitivity: ", round(cm$byClass["Sensitivity"], 4), "\n")
cat("Specificity: ", round(cm$byClass["Specificity"], 4), "\n")
```

Despite achieving 86% accuracy, the model correctly identifies only
\~16% of diabetes cases (sensitivity = 0.156). This is the core problem
with using accuracy under class imbalance — the model learns to predict
the majority class almost exclusively, which looks good on paper but is
clinically useless for identifying at-risk patients.

## Section 3b — Class Weights + 5-Fold Cross Validation

## Overview

To address class imbalance, we assign higher penalty to misclassifying
the minority class (diabetes) by weighting each observation inversely
proportional to its class frequency. This tells the model to treat one
minority class misclassification as more costly than a majority class
misclassification.

We evaluate using 5-fold cross validation rather than training error,
which gives an honest estimate of how the model generalizes to unseen
data — the F1 score is computed on held-out folds only, so it reflects
generalization performance, not memorization. We use **F1 score** as our
primary metric rather than accuracy — F1 balances precision and recall,
making it sensitive to how well we identify the minority class.

The test set is not touched here — CV is performed entirely within the
training set.

```{r class-weights-cv}
set.seed(571)

# Compute inverse frequency weights from training data
class_counts  <- table(binary_train$Diabetes_binary)
class_weights <- ifelse(binary_train$Diabetes_binary == "X1",
                        as.numeric(class_counts["X0"] / class_counts["X1"]),
                        1.0)

cat("Class counts:\n")
print(class_counts)
cat("Weight applied to X0:", 1.0, "\n")
cat("Weight applied to X1:", round(class_counts["X0"] / class_counts["X1"], 4), "\n")

# 5-fold CV setup
cv_ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# Train with class weights
model_weights <- train(
  Diabetes_binary ~ .,
  data      = binary_train,
  method    = "glm",
  family    = binomial,
  trControl = cv_ctrl,
  metric    = "ROC",
  weights   = class_weights
)

# CV held-out predictions
cv_preds       <- model_weights$pred
cv_preds$pred  <- factor(cv_preds$pred, levels = c("X0", "X1"))
cv_preds$obs   <- factor(cv_preds$obs,  levels = c("X0", "X1"))

cm_weights <- confusionMatrix(cv_preds$pred, cv_preds$obs, positive = "X1")

# Report
cat("\nCV ROC, Sensitivity, Specificity:\n")
print(round(model_weights$results[, c("ROC", "Sens", "Spec")], 4))

cat("\nCV Confusion Matrix:\n")
print(cm_weights$table)

tp <- cm_weights$table["X1", "X1"]
fp <- cm_weights$table["X1", "X0"]
fn <- cm_weights$table["X0", "X1"]

precision <- tp / (tp + fp)
recall    <- tp / (tp + fn)
f1        <- 2 * precision * recall / (precision + recall)

cat("\nPrecision:", round(precision, 4), "\n")
cat("Recall:   ", round(recall,    4), "\n")
cat("F1:       ", round(f1,        4), "\n")
```

Class weighting improves recall to 0.76, meaning the model correctly
identifies 76% of diabetes cases — a major improvement over the
baseline's 16%. However, precision drops to 0.31, indicating the model
over-predicts the minority class due to the aggressive 6.17x weight. F1
= 0.44 reflects this tradeoff. ROC-AUC of 0.82 indicates strong overall
discriminative ability — the model ranks diabetic cases above healthy
ones reliably, even when the classification threshold produces many
false positives.

## Section 3c — SMOTE Inside 5-Fold Cross Validation

## Overview

Instead of reweighting observations, SMOTE (Synthetic Minority
Oversampling Technique) generates synthetic minority class samples by
interpolating between existing minority class observations. This
balances the class distribution without simply repeating existing
samples.

Critically, SMOTE is applied **only inside each CV training fold** —
never on the full training set before CV. Applying SMOTE before CV would
cause data leakage because synthetic samples generated from the full
training set would be related to the validation fold observations,
producing overly optimistic estimates.

```{r smote-cv}
set.seed(571)

# Recipe defines preprocessing steps applied inside each CV fold
smote_recipe <- recipe(Diabetes_binary ~ ., data = binary_train) |>
  step_smote(Diabetes_binary, over_ratio = 1)  # balance to 1:1 ratio

# 5-fold CV setup — identical to 3b for fair comparison
cv_ctrl_smote <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# Train with SMOTE applied inside each fold via recipe
model_smote <- train(
  smote_recipe,
  data      = binary_train,
  method    = "glm",
  family    = binomial,
  trControl = cv_ctrl_smote,
  metric    = "ROC"
)

# CV held-out predictions
cv_preds_smote       <- model_smote$pred
cv_preds_smote$pred  <- factor(cv_preds_smote$pred, levels = c("X0", "X1"))
cv_preds_smote$obs   <- factor(cv_preds_smote$obs,  levels = c("X0", "X1"))

cm_smote <- confusionMatrix(cv_preds_smote$pred, cv_preds_smote$obs, positive = "X1")

# Report
cat("CV ROC, Sensitivity, Specificity:\n")
print(round(model_smote$results[, c("ROC", "Sens", "Spec")], 4))

cat("\nCV Confusion Matrix:\n")
print(cm_smote$table)

tp <- cm_smote$table["X1", "X1"]
fp <- cm_smote$table["X1", "X0"]
fn <- cm_smote$table["X0", "X1"]

precision <- tp / (tp + fp)
recall    <- tp / (tp + fn)
f1        <- 2 * precision * recall / (precision + recall)

cat("\nPrecision:", round(precision, 4), "\n")
cat("Recall:   ", round(recall,    4), "\n")
cat("F1:       ", round(f1,        4), "\n")

# Direct comparison with 3b
cat("\n--- Comparison: Class Weights vs SMOTE ---\n")
cat("         ROC     F1\n")
cat("Weights:", round(model_weights$results$ROC, 4),
                round(f1_weights <- {
                  tp_w <- cm_weights$table["X1","X1"]
                  fp_w <- cm_weights$table["X1","X0"]
                  fn_w <- cm_weights$table["X0","X1"]
                  p_w  <- tp_w/(tp_w+fp_w)
                  r_w  <- tp_w/(tp_w+fn_w)
                  2*p_w*r_w/(p_w+r_w)
                }, 4), "\n")
cat("SMOTE:  ", round(model_smote$results$ROC, 4), round(f1, 4), "\n")
```

SMOTE performs nearly identically to class weighting — F1 of 0.44 vs
0.44 and ROC-AUC of 0.82 vs 0.82. Both approaches produce similar
precision (0.31) and recall (0.76), meaning neither method has a
meaningful advantage for this dataset. This is not unusual with large,
well-structured survey data where the decision boundary is relatively
stable regardless of how imbalance is handled. We proceed to L2
regularization in 3d to determine whether penalizing large coefficients
improves performance further.

## Section 3d — L2 Regularized Logistic Regression (Ridge)

## Overview

Ridge regression (L2) adds a penalty proportional to the squared
magnitude of coefficients, shrinking all coefficients toward zero
without eliminating any. We prefer Ridge over Lasso here for two
reasons: first, all 21 features are theoretically motivated health
indicators with no reason to exclude any completely, which is what Lasso
does by driving coefficients to exactly zero. Second, correlated
predictors like `HighBP` and `HeartDiseaseorAttack` are handled more
stably by Ridge — Lasso tends to arbitrarily zero out one of a
correlated pair.

We use a two-step approach: first, `cv.glmnet` selects the optimal
lambda via internal CV. Then we pass that lambda to `caret` to obtain
true out-of-fold CV predictions — keeping evaluation consistent and
comparable with 3b and 3c.

```{r ridge-cv}
set.seed(571)

# Step 1 — use cv.glmnet to select optimal lambda
X_train <- model.matrix(Diabetes_binary ~ ., data = binary_train)[, -1]
y_train <- binary_train$Diabetes_binary

class_counts  <- table(binary_train$Diabetes_binary)
class_weights <- ifelse(y_train == "X1",
                        as.numeric(class_counts["X0"] / class_counts["X1"]),
                        1.0)

ridge_cv <- cv.glmnet(
  x            = X_train,
  y            = y_train,
  alpha        = 0,
  family       = "binomial",
  weights      = class_weights,
  nfolds       = 5,
  type.measure = "auc"
)

plot(ridge_cv, main = "Ridge — Cross Validated AUC vs Lambda")

cat("Best lambda (lambda.min):", round(ridge_cv$lambda.min, 6), "\n")
cat("Best lambda (lambda.1se):", round(ridge_cv$lambda.1se, 6), "\n")
cat("CV AUC at lambda.min:    ", round(max(ridge_cv$cvm), 4), "\n")

# Step 2 — use caret with the selected lambda for true out-of-fold CV evaluation
# cv.glmnet CV was for lambda selection only — not for final evaluation
# predicting on X_train directly would be in-sample and overly optimistic
cv_ctrl_ridge <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

model_ridge <- train(
  Diabetes_binary ~ .,
  data      = binary_train,
  method    = "glmnet",
  family    = "binomial",
  trControl = cv_ctrl_ridge,
  metric    = "ROC",
  weights   = class_weights,
  tuneGrid  = expand.grid(
    alpha  = 0,
    lambda = ridge_cv$lambda.min
  )
)

# True out-of-fold CV predictions
cv_preds_ridge       <- model_ridge$pred
cv_preds_ridge$pred  <- factor(cv_preds_ridge$pred, levels = c("X0", "X1"))
cv_preds_ridge$obs   <- factor(cv_preds_ridge$obs,  levels = c("X0", "X1"))

cm_ridge <- confusionMatrix(cv_preds_ridge$pred, cv_preds_ridge$obs, positive = "X1")

cat("\nCV ROC, Sensitivity, Specificity:\n")
print(round(model_ridge$results[, c("ROC", "Sens", "Spec")], 4))

cat("\nCV Confusion Matrix:\n")
print(cm_ridge$table)

tp <- cm_ridge$table["X1", "X1"]
fp <- cm_ridge$table["X1", "X0"]
fn <- cm_ridge$table["X0", "X1"]

precision <- tp / (tp + fp)
recall    <- tp / (tp + fn)
f1        <- 2 * precision * recall / (precision + recall)

cat("\nPrecision:", round(precision, 4), "\n")
cat("Recall:   ", round(recall,    4), "\n")
cat("F1:       ", round(f1,        4), "\n")
```

Ridge regularization produces nearly identical results to class
weighting and SMOTE — F1 of 0.44, ROC-AUC of 0.82, Precision of 0.31,
and Recall of 0.76. The flat CV AUC curve across lambda values confirms
that regularization provides no meaningful benefit here. With \~200k
observations and only 21 predictors, the model estimates stable
coefficients without penalization. Importantly, the confusion matrix
here is computed from true out-of-fold CV predictions via `caret` —
`cv.glmnet` was used solely to select lambda, not for evaluation.

All three approaches converge to the same performance, suggesting the
ceiling for logistic regression on this dataset is approximately F1 =
0.44 and ROC-AUC = 0.82.

## Comparison Summary

| Model          | ROC   | F1     | Precision | Recall |
|----------------|-------|--------|-----------|--------|
| Class Weights  | 0.8222| 0.4425 | 0.3113    | 0.7647 |
| SMOTE          | 0.8199| 0.4412 | 0.3109    | 0.7597 |
| Ridge          | 0.8220| 0.4421 | 0.3111    | 0.7639 |

## Section 3e — Model Selection and Retraining on Full Training Data

## Overview

All three approaches — class weights, SMOTE, and Ridge — produced nearly
identical CV performance (F1 ≈ 0.44, ROC-AUC ≈ 0.82). We select **class
weights** as the final approach based on marginal F1 advantage and
simplicity — it requires no synthetic data generation and no additional
regularization hyperparameter.

We retrain the selected model on the **full training set** with no CV
folds. This is standard practice — CV was used purely for honest
performance estimation. The final model should use all available
training data to learn the best possible coefficients before evaluating
on the test set once in 3f.

```{r retrain-final}
set.seed(571)

# Recompute class weights on full training set
class_counts        <- table(binary_train$Diabetes_binary)
class_weights_final <- ifelse(binary_train$Diabetes_binary == "X1",
                              as.numeric(class_counts["X0"] / class_counts["X1"]),
                              1.0)

# Retrain on full training set — no CV
final_model <- glm(
  Diabetes_binary ~ .,
  data    = binary_train,
  family  = binomial,
  weights = class_weights_final
)

cat("Final model trained on", nrow(binary_train), "observations\n")
cat("Number of coefficients:", length(coef(final_model)), "\n")

# Confirm no NA coefficients — would indicate perfect separation or collinearity
na_coefs <- sum(is.na(coef(final_model)))
cat("NA coefficients:", na_coefs, "\n")
```

We now have a single final model trained on all available training data.
This model is used exactly once in Section 3f to evaluate on the
held-out test set. CV estimates (F1 = 0.44, ROC-AUC = 0.82) are our
honest performance expectation — test set results should be close to
these if the model generalizes well.

```{r pre-test-check}
# Confirm final model exists and is clean
cat("Model class:", class(final_model), "\n")
cat("Coefficients:", length(coef(final_model)), "\n")
cat("NA coefficients:", sum(is.na(coef(final_model))), "\n")

# Confirm test set has not been modified beyond scaling
cat("\nbinary_test dimensions:", nrow(binary_test), "x", ncol(binary_test), "\n")
cat("Target is factor:", is.factor(binary_test$Diabetes_binary), "\n")
cat("Target levels:", levels(binary_test$Diabetes_binary), "\n")

# Confirm scaled columns look right
cat("\nTest set scaling check:\n")
print(round(sapply(binary_test[, c("BMI","MentHlth","PhysHlth")], 
      function(x) c(mean=mean(x), sd=sd(x))), 4))
```

## Section 3f — Final Evaluation on Test Set

## Overview

We evaluate the final model on the held-out test set exactly once. This
is the only time the test set is used in the entire binary
classification pipeline. The test set has never been used for training,
CV, preprocessing decisions, or threshold selection — so these results
represent a true estimate of how the model performs on unseen data.

We report a full confusion matrix, classification metrics, ROC-AUC
curve, and an odds ratio plot showing which features most strongly
predict diabetes risk.

```{r final-evaluation}
# Final predictions on test set
test_probs <- predict(final_model, newdata = binary_test, type = "response")
test_preds <- factor(ifelse(test_probs > 0.5, "X1", "X0"), levels = c("X0", "X1"))

# Confusion matrix
cm_test <- confusionMatrix(test_preds, binary_test$Diabetes_binary, positive = "X1")

cat("Test Set Confusion Matrix:\n")
print(cm_test$table)

tp <- cm_test$table["X1", "X1"]
fp <- cm_test$table["X1", "X0"]
fn <- cm_test$table["X0", "X1"]
tn <- cm_test$table["X0", "X0"]

precision <- tp / (tp + fp)
recall    <- tp / (tp + fn)
f1        <- 2 * precision * recall / (precision + recall)
accuracy  <- (tp + tn) / (tp + tn + fp + fn)

cat("\nAccuracy: ", round(accuracy,  4), "\n")
cat("Precision:", round(precision, 4), "\n")
cat("Recall:   ", round(recall,    4), "\n")
cat("F1:       ", round(f1,        4), "\n")
```

```{r roc-curve}
# ROC-AUC
roc_obj <- roc(
  response  = binary_test$Diabetes_binary,
  predictor = test_probs,
  levels    = c("X0", "X1"),
  direction = "<"
)

auc_val <- auc(roc_obj)
cat("Test ROC-AUC:", round(auc_val, 4), "\n")

# Plot ROC curve
roc_df <- data.frame(
  specificity = roc_obj$specificities,
  sensitivity = roc_obj$sensitivities
)

ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#2c7bb6", linewidth = 1) +
  geom_abline(linetype = "dashed", color = "gray50") +
  annotate("text", x = 0.75, y = 0.25,
           label = paste("AUC =", round(auc_val, 4)),
           size = 5, color = "#2c7bb6") +
  labs(
    title = "ROC Curve — Binary Logistic Regression (Test Set)",
    x     = "False Positive Rate",
    y     = "True Positive Rate"
  ) +
  theme_minimal()
```

```{r odds-ratio-plot}
# Odds ratios — exponentiated coefficients with 95% CIs
coef_summary <- summary(final_model)$coefficients
coef_df <- data.frame(
  Feature  = rownames(coef_summary),
  Estimate = coef_summary[, "Estimate"],
  SE       = coef_summary[, "Std. Error"]
) |>
  filter(Feature != "(Intercept)") |>
  mutate(
    OR      = exp(Estimate),
    CI_low  = exp(Estimate - 1.96 * SE),
    CI_high = exp(Estimate + 1.96 * SE)
  ) |>
  arrange(desc(OR))

# Print table so results are readable
cat("Odds Ratios (sorted by effect size):\n")
cat("OR > 1 = increases diabetes risk\n")
cat("OR < 1 = decreases diabetes risk\n")
cat("CI that does not cross 1 = statistically significant\n\n")
print(data.frame(
  Feature  = coef_df$Feature,
  OR       = round(coef_df$OR,      3),
  CI_low   = round(coef_df$CI_low,  3),
  CI_high  = round(coef_df$CI_high, 3)
))

# Visual plot
ggplot(coef_df, aes(x = OR, y = reorder(Feature, OR))) +
  geom_point(size = 3, color = "#2c7bb6") +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high),
                 height = 0.3, color = "#2c7bb6") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  labs(
    title = "Odds Ratios with 95% Confidence Intervals",
    x     = "Odds Ratio (exponentiated coefficient)",
    y     = NULL
  ) +
  theme_minimal()
```

The final model achieves a test ROC-AUC of 0.82, consistent with CV
estimates (0.82), confirming the model generalizes well to unseen data
with no signs of overfitting. Recall on the test set is 0.77, meaning
the model correctly identifies 77% of true diabetes cases — closely
matching the CV recall of 0.76. Precision remains low at 0.31,
reflecting the intentional tradeoff introduced by class weighting: the
model is tuned to catch as many true cases as possible at the cost of
more false alarms. In a health screening context, missing a true
diabetes case is more costly than incorrectly flagging a healthy
individual, so this tradeoff is acceptable.

The odds ratio table reveals which lifestyle factors most strongly
predict diabetes risk:

**Strongest risk factors (OR \> 1, CI does not cross 1):**

-   **CholCheck (OR = 3.65)** — the strongest predictor. Individuals who
    have had their cholesterol checked have 3.65x higher odds of
    diabetes, likely reflecting that people with known health conditions
    are more likely to seek cholesterol screening
-   **HighBP (OR = 2.08)** — high blood pressure more than doubles the
    odds of diabetes, consistent with their shared metabolic origins
-   **HighChol (OR = 1.80)** and **GenHlth (OR = 1.79)** — high
    cholesterol and poor general health are nearly equal predictors,
    each increasing odds by \~80%
-   **BMI (OR = 1.62)** — a one standard deviation increase in BMI
    increases odds of diabetes by 62%, confirming BMI as a strong
    modifiable risk factor
-   **Age (OR = 1.16)** — each step up the age scale increases odds by
    16%, reflecting the well-established relationship between age and
    diabetes risk

**Strongest protective factors (OR \< 1, CI does not cross 1):**

-   **HvyAlcoholConsump (OR = 0.48)** — heavy alcohol consumption is
    associated with lower diabetes odds. This is a known paradox in the
    literature and likely reflects survivorship bias or underreporting
    rather than a true protective effect
-   **PhysHlth (OR = 0.93)**, **Income (OR = 0.94)**, and **PhysActivity
    (OR = 0.96)** — better physical health, higher income, and physical
    activity are all associated with reduced diabetes risk

**Not statistically significant (CI crosses 1):**

-   **Smoker (OR = 1.00)** — no meaningful independent effect on
    diabetes odds after accounting for other features




# Section 4 — Multiclass Logistic Regression

## Overview

We extend the binary classification pipeline to a three-class problem:
no diabetes (X0), prediabetes (X1), and diabetes (X2). The class
distribution is severely imbalanced — prediabetes represents only 1.82%
of training observations, making this a harder problem than the binary
case. We use the same imbalance handling strategies as Section 3 and
report macro F1 as the primary metric, which treats all three classes
equally regardless of size.




## Section 4a — One-vs-Rest (OvR) with Class Weights + 5-Fold CV

## Overview

One-vs-Rest fits one binary logistic regression per class — each
classifier learns to distinguish one class from all others combined.
Final predictions are made by taking the class with the highest
predicted probability across all three classifiers. Class weights are
computed separately for each binary subproblem using inverse frequency
weighting.

Predictions are taken from held-out CV folds only so results are
honest and directly comparable to 4b.

```{r ovr-weights}
set.seed(571)

classes      <- c("X0", "X1", "X2")
ovr_models   <- list()
ovr_cv_preds <- list()

for (cls in classes) {
  cat("Training OvR classifier for class:", cls, "\n")
  
  # Binary target for this subproblem
  train_ovr        <- multiclass_train
  train_ovr$target <- factor(
    ifelse(multiclass_train$Diabetes_012 == cls, cls, "rest"),
    levels = c("rest", cls)
  )
  
  # Inverse frequency weights for this subproblem
  counts  <- table(train_ovr$target)
  weights <- ifelse(train_ovr$target == cls,
                    as.numeric(counts["rest"] / counts[cls]),
                    1.0)
  
  cat("Class:", cls, "| Count:", counts[cls],
      "| Weight:", round(counts["rest"] / counts[cls], 4), "\n")
  
  # 5-fold CV with held-out predictions
  cv_ctrl_ovr <- trainControl(
    method          = "cv",
    number          = 5,
    classProbs      = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final"
  )
  
  ovr_models[[cls]] <- train(
    target ~ . - Diabetes_012,
    data      = train_ovr,
    method    = "glm",
    family    = binomial,
    trControl = cv_ctrl_ovr,
    metric    = "ROC",
    weights   = weights
  )
  
  # Store held-out CV predictions
  ovr_cv_preds[[cls]] <- ovr_models[[cls]]$pred
}

# Combine OvR probabilities from held-out CV predictions only
n_train        <- nrow(multiclass_train)
combined_probs <- matrix(NA, nrow = n_train, ncol = length(classes))
colnames(combined_probs) <- classes

for (i in seq_along(classes)) {
  cls     <- classes[i]
  cv_pred <- ovr_cv_preds[[cls]]
  combined_probs[cv_pred$rowIndex, i] <- cv_pred[[cls]]
}

# Predict class with highest held-out probability
ovr_final_preds <- factor(
  classes[apply(combined_probs, 1, which.max)],
  levels = classes
)

# Remove any rows with NA probabilities
valid_rows  <- complete.cases(combined_probs)
ovr_preds_v <- ovr_final_preds[valid_rows]
ovr_obs_v   <- multiclass_train$Diabetes_012[valid_rows]

# Confusion matrix from held-out CV predictions
cm_ovr <- confusionMatrix(ovr_preds_v, ovr_obs_v)

cat("\nOvR CV Confusion Matrix (held-out predictions):\n")
print(cm_ovr$table)

cat("\nPer-class metrics:\n")
print(round(cm_ovr$byClass[, c("Precision", "Recall", "F1")], 4))

macro_f1_ovr <- mean(cm_ovr$byClass[, "F1"], na.rm = TRUE)
cat("\nMacro F1 (OvR):", round(macro_f1_ovr, 4), "\n")
```



One-vs-Rest with class weights produces a macro F1 of 0.44, driven
almost entirely by the no-diabetes (X0) and diabetes (X2) classifiers —
F1 of 0.81 and 0.45 respectively. The prediabetes classifier (X1)
achieves an F1 of only 0.06 despite a 53.95x class weight, correctly
identifying only 660 of 3,693 prediabetes cases. This is not a modeling
failure — it reflects the fundamental difficulty of distinguishing
prediabetes from the other two classes using survey-based lifestyle
features alone. With only 1.82% of observations being prediabetes, even
extreme penalization cannot compensate for the lack of representative
training signal. These results are now computed from true held-out CV
predictions and are directly comparable to 4b.


## Section 4b — Multinomial Logistic Regression via glmnet

## Overview

We use `glmnet` with `family = "multinomial"` and `alpha = 0` (Ridge)
as a faster and more scalable alternative to `nnet::multinom`. The math
is identical — multinomial logistic regression modeling all three classes
simultaneously — but `glmnet` uses coordinate descent optimization which
handles 200k+ observations efficiently. We use the same two-step approach
as Section 3d: `cv.glmnet` selects optimal lambda, then `caret` provides
true out-of-fold CV predictions for honest evaluation.

SMOTE is applied inside each CV fold via recipe, same as Section 3c,
to address the severe 1.82% prediabetes imbalance.

```{r multinom-glmnet}
set.seed(571)

# Step 1 — cv.glmnet to select optimal lambda
X_train_mc <- model.matrix(Diabetes_012 ~ ., data = multiclass_train)[, -1]
y_train_mc  <- multiclass_train$Diabetes_012

# Inverse frequency weights
counts_mc  <- table(y_train_mc)
max_count  <- max(counts_mc)
weights_mc <- as.numeric(max_count / counts_mc[y_train_mc])

cat("Class weights:\n")
print(round(max_count / counts_mc, 4))

ridge_cv_mc <- cv.glmnet(
  x            = X_train_mc,
  y            = y_train_mc,
  alpha        = 0,
  family       = "multinomial",
  weights      = weights_mc,
  nfolds       = 5,
  type.measure = "class"
)

plot(ridge_cv_mc, main = "Multinomial Ridge — CV Error vs Lambda")

cat("Best lambda (lambda.min):", round(ridge_cv_mc$lambda.min, 6), "\n")
cat("Best lambda (lambda.1se):", round(ridge_cv_mc$lambda.1se, 6), "\n")

# Step 2 — caret with selected lambda for true out-of-fold evaluation
mc_recipe <- recipe(Diabetes_012 ~ ., data = multiclass_train) |>
  step_smote(Diabetes_012, over_ratio = 0.5)

cv_ctrl_mc <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final"
)

model_multinom <- train(
  mc_recipe,
  data      = multiclass_train,
  method    = "glmnet",
  trControl = cv_ctrl_mc,
  metric    = "Mean_F1",
  tuneGrid  = expand.grid(
    alpha  = 0,
    lambda = ridge_cv_mc$lambda.min
  )
)

# CV held-out predictions
cv_preds_mc      <- model_multinom$pred
cv_preds_mc$pred <- factor(cv_preds_mc$pred, levels = classes)
cv_preds_mc$obs  <- factor(cv_preds_mc$obs,  levels = classes)

cm_multinom <- confusionMatrix(cv_preds_mc$pred, cv_preds_mc$obs)

cat("\nMultinomial CV Confusion Matrix:\n")
print(cm_multinom$table)

cat("\nPer-class metrics:\n")
print(round(cm_multinom$byClass[, c("Precision", "Recall", "F1")], 4))

macro_f1_mc     <- mean(cm_multinom$byClass[, "F1"], na.rm = TRUE)
class_counts_mc <- table(multiclass_train$Diabetes_012)
weighted_f1_mc  <- sum(
  cm_multinom$byClass[, "F1"] * as.numeric(class_counts_mc) /
  sum(class_counts_mc), na.rm = TRUE
)

cat("\nMacro F1:   ", round(macro_f1_mc,   4), "\n")
cat("Weighted F1:", round(weighted_f1_mc, 4), "\n")
```

Multinomial logistic regression via glmnet models all three classes
simultaneously using a single model, allowing it to learn shared
structure across classes rather than treating each as an independent
binary problem. The prediabetes class weight of 46.30x reflects the
severe 1.82% imbalance — the most extreme weighting in this entire
pipeline.

Despite this, prediabetes F1 remains low at 0.06 — only marginally
better than OvR's 0.06. This is a consistent finding across both
approaches and confirms that prediabetes is genuinely difficult to
distinguish from no-diabetes and diabetes using lifestyle survey
features alone. The two classes that bookend prediabetes — no diabetes
and diabetes — have overlapping feature profiles with prediabetes,
making separation inherently hard for a linear model.

Comparing the two approaches:

| Model       | Macro F1 | X0 F1 | X1 F1 | X2 F1 | Weighted F1 |
|-------------|----------|-------|-------|-------|-------------|
| OvR Weights | 0.4379   | 0.8074| 0.0583| 0.4479| —           |
| Multinomial | 0.4595   | 0.8765| 0.0631| 0.4390| 0.8007      |

Multinomial slightly outperforms OvR on macro F1 (0.46 vs 0.44) and
substantially improves no-diabetes F1 (0.88 vs 0.81), suggesting the
joint modeling of all three classes helps the majority class boundary.
We select multinomial as the final approach for 4c based on higher
macro F1.

## Section 4c — Model Selection and Retraining on Full Multiclass Training Data

## Overview

We select the best approach from 4a and 4b based on macro F1, then
retrain on the full multiclass training set before final evaluation.

```{r retrain-multiclass}
set.seed(571)

# Retrain final model using glmnet with best lambda from cv.glmnet
# No CV here — use all available training data
final_model_mc <- glmnet(
  x        = X_train_mc,
  y        = y_train_mc,
  alpha    = 0,
  family   = "multinomial",
  weights  = weights_mc,
  lambda   = ridge_cv_mc$lambda.min
)

cat("Final multiclass model trained on", nrow(multiclass_train), "observations\n")
cat("Lambda used:", round(ridge_cv_mc$lambda.min, 6), "\n")

# Confirm model has coefficients for all three classes
cat("Classes in model:", names(coef(final_model_mc)), "\n")
```

## Section 4d — Final Evaluation on Multiclass Test Set

## Overview

We evaluate the final multiclass model on the held-out test set exactly
once. We report a confusion matrix heatmap, per-class metrics, macro
and weighted F1, multiclass ROC-AUC using one-vs-rest, and a
coefficient plot showing odds ratios per class.


```{r multiclass-final-eval}
# glmnet requires matrix input
X_test_mc <- model.matrix(Diabetes_012 ~ ., data = multiclass_test)[, -1]

# Final predictions on test set
test_probs_mc <- predict(final_model_mc, newx = X_test_mc,
                         s = ridge_cv_mc$lambda.min, type = "response")

# test_probs_mc is a 3D array [obs x classes x 1] — drop third dimension
test_probs_mc <- test_probs_mc[, , 1]
colnames(test_probs_mc) <- classes

# Predicted class = highest probability
test_preds_mc <- factor(classes[apply(test_probs_mc, 1, which.max)],
                        levels = classes)

cm_test_mc <- confusionMatrix(test_preds_mc, multiclass_test$Diabetes_012)

cat("Test Set Confusion Matrix:\n")
print(cm_test_mc$table)

cat("\nPer-class metrics:\n")
print(round(cm_test_mc$byClass[, c("Precision", "Recall", "F1")], 4))

macro_f1_test     <- mean(cm_test_mc$byClass[, "F1"], na.rm = TRUE)
class_counts_test <- table(multiclass_test$Diabetes_012)
weighted_f1_test  <- sum(
  cm_test_mc$byClass[, "F1"] * as.numeric(class_counts_test) /
  sum(class_counts_test), na.rm = TRUE
)

cat("\nMacro F1:   ", round(macro_f1_test,   4), "\n")
cat("Weighted F1:", round(weighted_f1_test, 4), "\n")
cat("Accuracy:   ", round(cm_test_mc$overall["Accuracy"], 4), "\n")
```

```{r confusion-heatmap}
cm_df <- as.data.frame(cm_test_mc$table)
colnames(cm_df) <- c("Predicted", "Actual", "Count")

cm_df <- cm_df |>
  group_by(Actual) |>
  mutate(RowPct = round(Count / sum(Count) * 100, 1)) |>
  ungroup()

# Printed table
cat("Confusion Matrix with Row Percentages:\n")
print(cm_df |> arrange(Actual, Predicted))

# Heatmap
ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(Count, "\n(", RowPct, "%)")), size = 4) +
  scale_fill_gradient(low = "#deebf7", high = "#2c7bb6") +
  labs(
    title = "Confusion Matrix Heatmap — Multiclass Test Set",
    x     = "Predicted Class",
    y     = "Actual Class"
  ) +
  theme_minimal()
```

```{r multiclass-roc}
roc_x0 <- roc(response  = ifelse(multiclass_test$Diabetes_012 == "X0", 1, 0),
               predictor = test_probs_mc[, "X0"], quiet = TRUE)
roc_x1 <- roc(response  = ifelse(multiclass_test$Diabetes_012 == "X1", 1, 0),
               predictor = test_probs_mc[, "X1"], quiet = TRUE)
roc_x2 <- roc(response  = ifelse(multiclass_test$Diabetes_012 == "X2", 1, 0),
               predictor = test_probs_mc[, "X2"], quiet = TRUE)

cat("ROC-AUC per class (one-vs-rest):\n")
cat("X0 (No Diabetes):", round(auc(roc_x0), 4), "\n")
cat("X1 (Prediabetes):", round(auc(roc_x1), 4), "\n")
cat("X2 (Diabetes):   ", round(auc(roc_x2), 4), "\n")
cat("Average AUC:     ", round(mean(c(auc(roc_x0),
                                      auc(roc_x1),
                                      auc(roc_x2))), 4), "\n")
```

```{r multiclass-coef-plot}
# glmnet coef() returns a list — one matrix per class
coef_list <- lapply(classes, function(cls) {
  m <- as.matrix(coef(final_model_mc, s = ridge_cv_mc$lambda.min)[[cls]])
  data.frame(
    Class   = cls,
    Feature = rownames(m),
    OR      = round(exp(m[, 1]), 4)
  )
})

coef_mc_df <- do.call(rbind, coef_list) |>
  filter(Feature != "(Intercept)") |>
  mutate(Class = recode(Class,
                        "X0" = "No Diabetes",
                        "X1" = "Prediabetes",
                        "X2" = "Diabetes"))

# Printed table
cat("\nOdds Ratios by Class:\n")
cat("OR > 1 = increases odds of this class\n")
cat("OR < 1 = decreases odds of this class\n\n")
print(coef_mc_df |> arrange(Class, desc(OR)))

# Plot
ggplot(coef_mc_df, aes(x = OR, y = reorder(Feature, OR), color = Class)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  facet_wrap(~ Class, scales = "free_x") +
  labs(
    title = "Odds Ratios by Class — Multinomial Logistic Regression",
    x     = "Odds Ratio",
    y     = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
The final multiclass model achieves a test accuracy of 65.0% and a 
macro F1 of 0.43 — but these aggregate numbers obscure a stark 
three-way split in performance across classes.

No-diabetes (X0) performs well with an F1 of 0.79, driven by high 
precision (0.95). The model is rarely wrong when it predicts X0, but 
misses 33% of true X0 cases, routing them to X1 or X2. Diabetes (X2) 
reaches F1 = 0.44 with recall of 0.58 — identifying just over half of 
true diabetes cases, at the cost of moderate false positives from the 
other classes.

Prediabetes (X1) is the clear failure point. Despite a 46x class 
weight, the model correctly identifies only 277 of 938 true prediabetes 
cases (recall = 0.30, F1 = 0.056). Of the remaining cases, 30.7% are 
predicted as no-diabetes and 39.8% as diabetes — the model distributes 
prediabetes predictions nearly uniformly across all three classes. This 
is not a tuning failure. Prediabetes occupies an intermediate feature 
space between no-diabetes and diabetes, and the flat coefficient profile 
for X1 (all ORs between 0.92–1.25, no dominant predictor) confirms 
there is no strong linear signal separating it from adjacent classes 
using lifestyle survey data alone.

ROC-AUC tells the same story: X0 and X2 achieve AUC of 0.81 and 0.82 
respectively, consistent with binary task performance. Prediabetes AUC 
of 0.68 — above chance but far below the other classes — confirms the 
model has weak but nonzero discriminative ability for X1 that it cannot 
translate into reliable classifications.

The diabetes coefficient plot mirrors Section 3f: CholCheck (OR = 1.69), 
HighBP (OR = 1.43), GenHlth (OR = 1.30), HighChol (OR = 1.24), and 
BMI (OR = 1.21) are the strongest predictors of diabetes — the same 
metabolic risk factors that dominated the binary task. HvyAlcoholConsump 
again appears as a negative predictor of diabetes (OR = 0.66), 
replicating the paradox noted in Section 3f. Test macro F1 of 0.43 
closely matches the CV estimate of 0.46, confirming the model 
generalizes without meaningful overfitting.






# Section 5 — Model Comparison

## Overview

We consolidate all model variants from Sections 3 and 4 into two
comparison tables — one per task. Throughout, class labels refer to:
**X0 = No Diabetes**, **X1 = Prediabetes**, **X2 = Diabetes**. CV F1
was the sole criterion for model selection — test metrics were computed
exactly once for the selected model in each task and are not available
for variants not carried forward. Selected models are highlighted in
green.

```{r binary-comparison-table}
library(kableExtra)

# --- Table 1: Binary Task ---
binary_df <- data.frame(
  Model      = c("Logistic Regression",
                 "Logistic Regression",
                 "Ridge Logistic (glmnet)"),
  Imbalance  = c("Class Weights", "SMOTE", "Class Weights"),
  CV_F1      = c(0.4425, 0.4412, 0.4421),
  CV_AUC     = c(0.8222, 0.8199, 0.8220),
  Test_Acc   = c("0.7310", "—", "—"),
  Test_Prec  = c("0.3116", "—", "—"),
  Test_Rec   = c("0.7692", "—", "—"),
  Test_F1    = c("0.4435", "—", "—"),
  Test_AUC   = c("0.8237", "—", "—")
)

binary_df |>
  kbl(
    col.names = c("Model", "Imbalance Method",
                  "CV F1", "CV ROC-AUC",
                  "Test Accuracy", "Test Precision",
                  "Test Recall", "Test F1", "Test ROC-AUC"),
    caption   = "Table 1. Binary Task — Model Comparison.
                 Test metrics reported only for the selected model.
                 — indicates variants not carried forward to test evaluation.",
    align     = "llcccccccc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE
  ) |>
  row_spec(1, background = "#d4edda") |>   # selected model
  column_spec(3, bold = TRUE) |>           # CV F1 — selection criterion
  column_spec(8, bold = TRUE)              # Test F1 — primary outcome
```

```{r multiclass-comparison-table}
# --- Table 2: Multiclass Task — Model Selection ---
mc_df <- data.frame(
  Model      = c("One-vs-Rest (glm)",
                 "Multinomial Ridge (glmnet)"),
  Imbalance  = c("Class Weights", "SMOTE"),
  CV_F1      = c(0.4379, 0.4595),
  Test_F1    = c("—", "0.4263 (macro)"),
  Test_AUC   = c("—", "0.7703 (avg)"),
  Accuracy   = c("—", "0.6504")
)

mc_df |>
  kbl(
    col.names = c("Model", "Imbalance Method",
                  "CV Macro F1", "Test Macro F1",
                  "Test ROC-AUC (avg)", "Test Accuracy"),
    caption   = "Table 2. Multiclass Task — Model Comparison.
                 Test metrics reported only for the selected model.
                 — indicates variant not carried forward to test evaluation.",
    align     = "llcccc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) |>
  row_spec(2, background = "#d4edda") |>   # selected model
  column_spec(3, bold = TRUE) |>           # CV F1 — selection criterion
  column_spec(4, bold = TRUE)              # Test F1 — primary outcome

# --- Table 3: Per-Class Test Performance (selected model only) ---
mc_class_df <- data.frame(
  Class     = c("X0 — No Diabetes",
                "X1 — Prediabetes",
                "X2 — Diabetes"),
  Precision = c(0.9471, 0.0310, 0.3527),
  Recall    = c(0.6704, 0.2953, 0.5766),
  F1        = c(0.7851, 0.0561, 0.4377),
  ROC_AUC   = c(0.8131, 0.6802, 0.8177)
)

mc_class_df |>
  kbl(
    col.names = c("Class", "Precision", "Recall", "F1", "ROC-AUC"),
    caption   = "Table 3. Multinomial Ridge — Per-Class Test Performance.",
    digits    = 4,
    align     = "lcccc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) |>
  row_spec(2, background = "#ffe8e8") |>   # X1 — highlight failure
  column_spec(4, bold = TRUE) |>           # F1
  column_spec(5, bold = TRUE)              # ROC-AUC
```

Four findings stand out across these tables.

**Imbalance method does not drive binary performance.** Class weights,
SMOTE, and Ridge regularization converge to CV F1 ≈ 0.44 and CV
ROC-AUC ≈ 0.82 across the board. With ~200k observations and 21
predictors, the decision boundary is stable regardless of how imbalance
is handled. The performance ceiling for logistic regression on this
binary task is approximately F1 = 0.44 and ROC-AUC = 0.82 — determined
by the model family and feature set, not the resampling strategy. Class
weights was selected as the binary winner on marginal CV F1 advantage
and simplicity.

**Multinomial Ridge was selected over One-vs-Rest** based on higher CV
macro F1 (0.46 vs 0.44). One-vs-Rest treats each class as an
independent binary problem and never sees the full three-class structure
during training. Multinomial Ridge models all three classes
simultaneously, allowing shared structure across class boundaries to
inform each coefficient. The test macro F1 of 0.43 coming in marginally
below CV (0.46) is within normal variance and does not indicate
overfitting.

**The binary task outperforms multiclass on every aggregate metric.**
Binary test ROC-AUC of 0.82 versus multiclass average AUC of 0.77;
binary recall of 0.77 versus multiclass X2 recall of 0.58. This gap is
driven entirely by X1 — X0 and X2 achieve AUC values of 0.81 and 0.82,
fully consistent with binary performance. Collapsing prediabetes into
the positive class in the binary task sidesteps the hardest boundary in
the data; the multiclass task forces the model to draw it explicitly and
cannot do so reliably with lifestyle survey features alone.

**Per-class metrics expose the prediabetes problem clearly.** X0
achieves high precision (0.95) but moderate recall (0.67). X2 reaches
balanced moderate performance (F1 = 0.44, AUC = 0.82). X1 fails on
every metric: precision 0.03, recall 0.30, F1 = 0.056, AUC = 0.68 —
highlighted in red above. No imbalance strategy can compensate for the
structural ambiguity of prediabetes in feature space — its lifestyle
profile overlaps substantially with both neighboring classes, and no
linear boundary can reliably isolate it.


# Section 6 — Key Findings & Interpretation

**The same metabolic risk factors dominate both tasks.** Across binary
and multiclass models, the strongest predictors of diabetes are
CholCheck, HighBP, HighChol, GenHlth, and BMI — an interconnected
cluster of metabolic conditions with shared physiological origins. Their
consistency across two modeling approaches and two target definitions
strengthens confidence these associations are structural, not artifacts
of model choice.

**Separating prediabetes from diabetes reveals the limits of this
feature set.** The binary task achieves ROC-AUC = 0.82 by collapsing
prediabetes and diabetes into one class. The multiclass task attempts
to draw that boundary explicitly and fails — X1 F1 = 0.056 despite a
46x class weight. Prediabetes and diabetes share nearly identical
lifestyle risk profiles; the difference between them is likely captured
by clinical biomarkers such as blood glucose and HbA1c that are absent
from this dataset.

**Heavy alcohol consumption appears protective — it is not.**
HvyAlcoholConsump shows OR = 0.48 (binary) and OR = 0.66 (multiclass
diabetes class). This is a well-documented paradox in epidemiological
survey data, likely driven by survivorship bias, underreporting, or the
healthy user effect rather than any true protective mechanism.

**Self-reported data introduces measurement error in both directions.**
BRFSS is a telephone survey — diabetes status, lifestyle behaviors, and
health conditions are all self-reported. This means both the features
and the labels carry measurement error simultaneously, which attenuates
coefficient estimates and suppresses model performance in ways that
cannot be corrected by modeling choices alone.

**Logistic regression has a hard ceiling on this problem.** F1 = 0.44
and ROC-AUC = 0.82 represent the limit of what a linear model can
extract from these features. Diabetes risk is driven by complex
interactions between BMI, age, blood pressure, and metabolic history
that a single linear combination cannot fully capture. Tree-based models
such as gradient boosting or random forests, which naturally model
feature interactions and nonlinear boundaries, would be the appropriate
next step.
