---
title: "Random Forest"
author: "Matthew"
date: "2026-04-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Install these packages first if you don't have them: 
# install.packages(c("ranger", "caret", "pROC", "dplyr", "ggplot2"))

library(ranger)
library(caret)
library(pROC)
library(dplyr)
library(ggplot2)

print("Pop-up 1: Please find and double-click 'binary_train.csv'")
binary_train <- read.csv(file.choose())

print("Pop-up 2: Please find and double-click 'multiclass_train.csv'")
multiclass_train <- read.csv(file.choose())

print("Pop-up 3: Please find and double-click 'binary_test.csv'")
binary_test <- read.csv(file.choose()) 

print("Pop-up 4: Please find and double-click 'multiclass_test.csv'")
multiclass_test <- read.csv(file.choose())

binary_train$Diabetes_binary <- as.factor(binary_train$Diabetes_binary)
binary_test$Diabetes_binary <- as.factor(binary_test$Diabetes_binary)

multiclass_train$Diabetes_012 <- as.factor(multiclass_train$Diabetes_012)
multiclass_test$Diabetes_012 <- as.factor(multiclass_test$Diabetes_012)

bin_counts <- table(binary_train$Diabetes_binary)
bin_weights_calc <- sum(bin_counts) / (2 * bin_counts)
bin_case_weights <- ifelse(binary_train$Diabetes_binary == "0", bin_weights_calc["0"], bin_weights_calc["1"])

multi_counts <- table(multiclass_train$Diabetes_012)
multi_weights_calc <- sum(multi_counts) / (3 * multi_counts)
multi_case_weights <- case_when(
  multiclass_train$Diabetes_012 == "0" ~ multi_weights_calc["0"],
  multiclass_train$Diabetes_012 == "1" ~ multi_weights_calc["1"],
  multiclass_train$Diabetes_012 == "2" ~ multi_weights_calc["2"]
)

print("Training Binary Random Forest...")
rf_binary <- ranger(
  dependent.variable.name = "Diabetes_binary",
  data = binary_train,
  num.trees = 500,           # Tuning: n_estimators
  max.depth = 15,            # Tuning: max_depth (prevents overfitting)
  importance = "impurity",   # Gini/Entropy importance
  case.weights = bin_case_weights, # Handles the class imbalance
  probability = TRUE         # Needed for ROC-AUC
)

print("Training Multiclass Random Forest...")
rf_multi <- ranger(
  dependent.variable.name = "Diabetes_012",
  data = multiclass_train,
  num.trees = 500,
  max.depth = 15,
  importance = "impurity",
  case.weights = multi_case_weights,
  probability = TRUE
)

preds_bin_prob <- predict(rf_binary, binary_test)$predictions
preds_bin_class <- as.factor(ifelse(preds_bin_prob[,2] > 0.5, "1", "0"))

preds_multi_prob <- predict(rf_multi, multiclass_test)$predictions
preds_multi_class <- as.factor(apply(preds_multi_prob, 1, which.max) - 1)

print("--- BINARY MODEL METRICS ---")
cm_bin <- confusionMatrix(preds_bin_class, binary_test$Diabetes_binary, mode = "everything")
print(cm_bin)

roc_obj <- roc(binary_test$Diabetes_binary, preds_bin_prob[,2])
print(paste("ROC-AUC:", round(auc(roc_obj), 4)))

print("--- MULTICLASS MODEL METRICS ---")
cm_multi <- confusionMatrix(preds_multi_class, multiclass_test$Diabetes_012, mode = "everything")
print(cm_multi)

imp_bin <- data.frame(Feature = names(rf_binary$variable.importance), 
                      Importance = rf_binary$variable.importance)
imp_multi <- data.frame(Feature = names(rf_multi$variable.importance), 
                        Importance = rf_multi$variable.importance)

ggplot(imp_bin, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Feature Importance (Gini): Binary Random Forest", x = "Features", y = "Importance")

ggplot(imp_multi, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "firebrick") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Feature Importance (Gini): Multiclass Random Forest", x = "Features", y = "Importance")

print("--- BIAS CHECK: RECALL BY AGE GROUP ---")
binary_test$Predicted <- preds_bin_class

bias_check <- binary_test %>%
  filter(Diabetes_binary == "1") %>% # Look only at actual positive cases
  group_by(Age) %>%
  summarize(
    Total_Actual_Diabetics = n(),
    Correctly_Predicted = sum(Predicted == "1"),
    Recall = Correctly_Predicted / Total_Actual_Diabetics
  )
print(bias_check)
```
  Metrics for the Master Table
Binary Model (Random Forest)

Accuracy: 0.7506

Precision (Class 1): 0.3265

Recall (Class 1): 0.7433

F1 Score (Class 1): 0.4536 (Calculated from your Precision/Recall)

ROC-AUC: 0.8279

  Multiclass Model (Random Forest)

Accuracy: 0.7138

Class 0 (No Diabetes) F1: 0.8209

Class 1 (Prediabetes) F1: 0.0428

Class 2 (Diabetes) F1: 0.4481

Macro F1: 0.4373 (The average of the three F1 scores above)



  Model Performance Metrics
The Binary Random Forest model achieved an overall accuracy of 0.7506, alongside a ROC-AUC of 0.8279. When evaluating its specific ability to predict actual diabetes cases, the model returned a precision of 0.3265 and a recall of 0.7433, resulting in a Class 1 F1 score of 0.4536. The Multiclass model yielded a slightly lower overall accuracy of 0.7138. A breakdown of the class-specific F1 scores highlights a significant structural challenge with the prediabetes category. While Class 0 (No Diabetes) and Class 2 (Diabetes) achieved F1 scores of 0.8209 and 0.4481 respectively, the Class 1 (Prediabetes) F1 score plummeted to just 0.0428, pulling the overall macro F1 score down to 0.4373. This perfectly validates Matthias's earlier hypothesis: the inability to predict prediabetes is a feature-set problem stemming from missing biomarkers, not a modeling or algorithmic limitation.

  Bias & Fairness Analysis
An analysis of the binary model's recall across the dataset's 13 age brackets reveals a severe demographic bias. The model performs exceptionally poorly on younger individuals, successfully identifying actual diabetics (recall) at a rate of only 0% to 16% for age brackets 1 through 3. Conversely, the model's recall improves drastically with age, peaking at over 81% for senior brackets 10 through 12. This indicates that the algorithm is heavily biased toward assuming younger people do not have diabetes, relying disproportionately on advanced age to trigger a positive prediction and consequently missing the vast majority of young diabetic patients in the test set.

  Feature Importance Comparison
Comparing the Gini impurity scores between the two datasets reveals a notable shift in how the models prioritize information. In the binary model, broad, subjective health indicators dominate the decision-making process, with General Health (GenHlth) and High Blood Pressure (HighBP) ranking as the top two most important features, followed by BMI and Age. However, in the multiclass model, raw physical metrics and baseline demographics take over. When the Random Forest is forced to separate three distinct classes—particularly the difficult prediabetes group, it relies overwhelmingly on the patient's BMI and Age as the primary drivers, pushing GenHlth and Income down to the third and fourth positions.
