---
title: "XGBoost"
author: "Zach Skiba"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Section 0 — Setup & Package Loading

### Overview

We load all required packages and set a global seed for reproducibility.
Key packages: `xgboost` for gradient boosted trees, `SHAPforxgboost` for
SHAP value computation and visualization, `caret` for confusion matrix
utilities, `pROC` for ROC-AUC curves, and `ggplot2` / `kableExtra` for
visualization and reporting.

The seed is set once here and repeated in any chunk that involves
randomness to ensure reproducibility across non-sequential chunk execution.


```{r load-packages}
required_packages <- c(
  "xgboost", "SHAPforxgboost", "caret", "pROC",
  "ggplot2", "kableExtra", "dplyr", "tidyr", "Matrix"
)

installed  <- rownames(installed.packages())
to_install <- setdiff(required_packages, installed)
if (length(to_install) > 0) install.packages(to_install, dependencies = TRUE)

library(xgboost)
library(SHAPforxgboost)
library(caret)
library(pROC)
library(ggplot2)
library(kableExtra)
library(dplyr)
library(tidyr)
library(Matrix)

set.seed(571)

cat("All packages loaded successfully.\n")
cat("xgboost version:", as.character(packageVersion("xgboost")), "\n")
```


## Section 1 — Data Loading & Validation

### Overview

We load all four pre-split CSVs directly — no re-splitting is done here.
The data was stratified 80/20 by a teammate. We verify dimensions, column
alignment, missing values, and class distributions for both targets.

Unlike the logistic regression notebook, we do NOT convert targets to
factors here — XGBoost requires numeric labels (0/1 for binary, 0/1/2
for multiclass).


```{r load-data}
binary_train     <- read.csv("../data/processed/binary_train.csv")
binary_test      <- read.csv("../data/processed/binary_test.csv")
multiclass_train <- read.csv("../data/processed/multiclass_train.csv")
multiclass_test  <- read.csv("../data/processed/multiclass_test.csv")
```

```{r validate}
# Dimensions
cat("binary_train:    ", nrow(binary_train),     "x", ncol(binary_train),     "\n")
cat("binary_test:     ", nrow(binary_test),      "x", ncol(binary_test),      "\n")
cat("multiclass_train:", nrow(multiclass_train), "x", ncol(multiclass_train), "\n")
cat("multiclass_test: ", nrow(multiclass_test),  "x", ncol(multiclass_test),  "\n")

# Column alignment
cat("\nColumns match (binary):    ",
    identical(colnames(binary_train), colnames(binary_test)), "\n")
cat("Columns match (multiclass):",
    identical(colnames(multiclass_train), colnames(multiclass_test)), "\n")

# Missing values
cat("\nNA counts:\n")
cat("binary_train:",     sum(is.na(binary_train)),     "\n")
cat("binary_test:",      sum(is.na(binary_test)),      "\n")
cat("multiclass_train:", sum(is.na(multiclass_train)), "\n")
cat("multiclass_test:",  sum(is.na(multiclass_test)),  "\n")
```

```{r class-distribution}
cat("Binary target distribution (train):\n")
print(round(prop.table(table(binary_train$Diabetes_binary)) * 100, 2))

cat("\nMulticlass target distribution (train):\n")
print(round(prop.table(table(multiclass_train$Diabetes_012)) * 100, 2))
```

The four datasets load cleanly with no missing values. Dimensions and
column alignment are confirmed across train and test sets. Class
imbalance is present in both tasks — 86/14 for binary and 84/2/14 for
multiclass — consistent with the EDA findings.

```{r preprocessing}
set.seed(571)

scale_cols <- c("BMI", "MentHlth", "PhysHlth")

train_means <- colMeans(binary_train[, scale_cols])
train_sds   <- apply(binary_train[, scale_cols], 2, sd)

binary_train[, scale_cols] <- scale(binary_train[, scale_cols],
                                    center = train_means, scale = train_sds)
binary_test[, scale_cols]  <- scale(binary_test[, scale_cols],
                                    center = train_means, scale = train_sds)

mc_means <- colMeans(multiclass_train[, scale_cols])
mc_sds   <- apply(multiclass_train[, scale_cols], 2, sd)

multiclass_train[, scale_cols] <- scale(multiclass_train[, scale_cols],
                                        center = mc_means, scale = mc_sds)
multiclass_test[, scale_cols]  <- scale(multiclass_test[, scale_cols],
                                        center = mc_means, scale = mc_sds)

spw <- sum(binary_train$Diabetes_binary == 0) / sum(binary_train$Diabetes_binary == 1)
cat("scale_pos_weight:", round(spw, 4), "\n")

train_idx <- createDataPartition(binary_train$Diabetes_binary, p = 0.9, list = FALSE)
tune_data  <- binary_train[ train_idx, ]
val_data   <- binary_train[-train_idx, ]

X_tune <- as.matrix(tune_data[, setdiff(names(tune_data), "Diabetes_binary")])
y_tune <- tune_data$Diabetes_binary
X_val  <- as.matrix(val_data[, setdiff(names(val_data),  "Diabetes_binary")])
y_val  <- val_data$Diabetes_binary

dtune <- xgb.DMatrix(data = X_tune, label = y_tune)
dval  <- xgb.DMatrix(data = X_val,  label = y_val)

X_train <- as.matrix(binary_train[, setdiff(names(binary_train), "Diabetes_binary")])
y_train <- binary_train$Diabetes_binary

cat("Training observations:", nrow(binary_train), "\n")
cat("Tuning slice:         ", nrow(tune_data), "\n")
cat("Validation slice:     ", nrow(val_data), "\n")
```

## Section 2 — Binary XGBoost

### Overview

We train an XGBoost classifier on the binary task (`Diabetes_binary`:
0 = no diabetes, 1 = prediabetes/diabetes). XGBoost is a gradient
boosted tree ensemble that builds trees sequentially, each one correcting
the residual errors of the previous. Unlike logistic regression, it
naturally captures nonlinear relationships and feature interactions
without any explicit specification.

**Imbalance handling:** We use `scale_pos_weight`, XGBoost's native
parameter for binary class imbalance. It is set to the ratio of negative
to positive class counts in the training data, which tells the loss
function to penalize minority class misclassifications proportionally
more. This is equivalent to class weighting in sklearn and is the
recommended approach for tree-based models — no synthetic data generation
is needed or appropriate here.

**Tuning strategy:** We use a stratified 90/10 validation slice carved
from the training set. XGBoost's `early_stopping_rounds` monitors val
AUC after every boosting round and halts when it stops improving,
automatically selecting the optimal number of trees. We then perform a
manual grid search over `max_depth` and `eta` (learning rate) — the two
parameters with the largest practical impact on XGBoost performance.
`subsample` and `colsample_bytree` are set to 0.8 throughout, which is a
well-established default that provides regularization through stochastic
sampling. After selecting the best hyperparameters, the final model is
retrained on the full training set before test evaluation.

**Why we do not use cross-validation:** XGBoost's early stopping on a
validation watchlist serves the same role as CV for simpler models —
it provides an honest generalization estimate and prevents overfitting
by stopping tree growth at the right point. Running full k-fold CV on
XGBoost at this data size would be computationally expensive without
meaningfully improving hyperparameter selection.


``` {r binary-grid-search-upgraded}

set.seed(571)

# 1. Expanded hyperparameter grid 
param_grid <- expand.grid(
  max_depth        = c(4, 6),
  eta              = c(0.03, 0.1),
  min_child_weight = c(1, 10),
  subsample        = c(0.6, 0.8),
  colsample_bytree = c(0.6, 0.8),
  gamma            = c(0, 0.1)
)

cat("Total combinations:", nrow(param_grid), "\n\n")

# 2. repeated evaluation seeds
seeds <- c(571, 123, 999)

results <- data.frame()

for (i in seq_len(nrow(param_grid))) {

  auc_vals <- c()

  for (s in seeds) {

    set.seed(s)

    params <- list(
      objective        = "binary:logistic",
      eval_metric      = "auc",
      scale_pos_weight = spw,
      max_depth        = param_grid$max_depth[i],
      eta              = param_grid$eta[i],
      min_child_weight = param_grid$min_child_weight[i],
      subsample        = param_grid$subsample[i],
      colsample_bytree = param_grid$colsample_bytree[i],
      gamma            = param_grid$gamma[i],
      seed             = s
    )

    model <- xgb.train(
      params  = params,
      data    = dtune,
      nrounds = 1000,
      evals   = list(val = dval),
      early_stopping_rounds = 30,
      verbose = 0
    )

    eval_log <- attributes(model)$evaluation_log

    if (!is.null(eval_log) && "val_auc" %in% colnames(eval_log)) {
      auc_vals <- c(auc_vals, max(eval_log$val_auc))
    }
  }

  # 3. aggregate performance
  mean_auc <- mean(auc_vals, na.rm = TRUE)
  auc_sd   <- sd(auc_vals, na.rm = TRUE)

  n <- length(auc_vals)
  ci_half_width <- 1.96 * auc_sd / sqrt(n)

  ci_lower <- mean_auc - ci_half_width
  ci_upper <- mean_auc + ci_half_width

  # 4. store results
  results <- rbind(results, data.frame(
    max_depth        = param_grid$max_depth[i],
    eta              = param_grid$eta[i],
    min_child_weight = param_grid$min_child_weight[i],
    subsample        = param_grid$subsample[i],
    colsample_bytree = param_grid$colsample_bytree[i],
    gamma            = param_grid$gamma[i],
    mean_auc         = round(mean_auc, 5),
    std_auc          = round(auc_sd, 5),
    ci_lower         = round(ci_lower, 5),
    ci_upper         = round(ci_upper, 5)
  ))

  # 5. logging
  cat(sprintf(
    "Combo %02d/%02d | depth=%d eta=%.2f mcw=%d subs=%.1f col=%.1f gamma=%.1f | AUC=%.5f ± %.5f | CI [%.5f, %.5f]\n",
    i, nrow(param_grid),
    param_grid$max_depth[i],
    param_grid$eta[i],
    param_grid$min_child_weight[i],
    param_grid$subsample[i],
    param_grid$colsample_bytree[i],
    param_grid$gamma[i],
    mean_auc,
    auc_sd,
    ci_lower,
    ci_upper
  ))
}

# 6. rank models robustly
results <- results[order(-results$mean_auc), ]

cat("\nTop 5 configs (robust mean AUC):\n")
print(head(results, 5))

best <- results[1, ]

cat("\nBest configuration:\n")
print(best)


```

The expanded grid search shows that max_depth = 4 consistently dominates, indicating that predictive signal in this dataset is primarily captured through shallow feature interactions rather than deep hierarchical structure. Across all 64 configurations, the AUC range is approximately 0.0012, confirming that performance is largely insensitive to hyperparameter variation within this regime.

The best-performing models form a tight cluster around AUC ≈ 0.8275–0.8276, with all differences well within overlapping confidence intervals. This suggests that no single configuration is meaningfully superior in a statistical sense, and observed variations are effectively noise.

Given this near-equivalence, we prioritize parsimony and stability over marginal gains. We select a representative configuration from the top-performing cluster:

max_depth = 4, eta = 0.10, min_child_weight = 10, subsample = 0.6, colsample_bytree = 0.6, gamma = 0.0

This choice remains within the highest-performing band while reducing model complexity and boosting efficiency (fewer trees under early stopping), without sacrificing predictive performance. The resulting model achieves AUC ≈ 0.8275 with a substantially simpler and more stable configuration.


### Section 2b — Final Binary Model Training on Full Training Set

After selecting the best hyperparameters from the validation search, we retrain the final
model on the complete binary training set with no held-out validation slice. This is standard
practice — the validation slice was used only for hyperparameter selection. The final model
should learn from all available training data before being evaluated on the test set once in
Section 2c.

```{r binary-final-train}
set.seed(571)

best_params <- list(
  objective        = "binary:logistic",
  eval_metric      = "auc",
  scale_pos_weight = spw,
  max_depth        = 4,
  eta              = 0.10,
  min_child_weight = 10,
  subsample        = 0.6,
  colsample_bytree = 0.6,
  gamma            = 0.0,
  seed             = 571
)

# full training matrix
dtrain_full <- xgb.DMatrix(
  data  = as.matrix(binary_train[, setdiff(names(binary_train), "Diabetes_binary")]),
  label = binary_train$Diabetes_binary
)

# final model (hardcoded optimal trees)
final_model_binary <- xgb.train(
  params  = best_params,
  data    = dtrain_full,
  nrounds = 167,
  verbose = 0
)

cat("Final binary model trained\n")
cat("Number of trees:", 167, "\n")
cat("Training observations:", nrow(binary_train), "\n")

```

### Section 2c — Final Evaluation on Binary Test Set

The final model is evaluated exactly once on `binary_test` — data that has never been
touched during training, validation slicing, or hyperparameter tuning. We report the full
confusion matrix, precision, recall, F1, and ROC-AUC to allow direct comparison against
the logistic regression baseline (F1 = 0.44, AUC = 0.82). The primary question is whether
XGBoost's capacity for nonlinear boundaries and feature interactions translates into a
meaningful lift over the linear baseline on this dataset.

Predictions are generated as probabilities via `predict()` and thresholded at 0.5. The
confusion matrix uses class 1 as the positive class. ROC-AUC is computed using `pROC`
and plotted with `ggplot2`. All metrics are computed on the test set exactly once and are
the final reported results for the binary task.

```{r binary-final-eval}
set.seed(571)

X_test <- as.matrix(binary_test[, setdiff(names(binary_test), "Diabetes_binary")])
y_test  <- binary_test$Diabetes_binary

dtest <- xgb.DMatrix(data = X_test, label = y_test)

probs <- predict(final_model_binary, dtest)
preds <- ifelse(probs > 0.5, 1, 0)

cm <- confusionMatrix(
  factor(preds, levels = c(0, 1)),
  factor(y_test, levels = c(0, 1)),
  positive = "1"
)

print(cm$table)

tp <- cm$table[2, 2]
fp <- cm$table[2, 1]
fn <- cm$table[1, 2]

precision <- tp / (tp + fp)
recall    <- tp / (tp + fn)
f1        <- 2 * precision * recall / (precision + recall)

cat("Accuracy: ",  round(cm$overall["Accuracy"], 4), "\n")
cat("Precision:",  round(precision, 4), "\n")
cat("Recall:   ",  round(recall,    4), "\n")
cat("F1:       ",  round(f1,        4), "\n")

roc_obj <- roc(y_test, probs)
cat("ROC-AUC:  ",  round(auc(roc_obj), 4), "\n")

ggroc(roc_obj, color = "#2c7bb6", linewidth = 1) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray50") +
  labs(
    title = "Binary XGBoost — ROC Curve (Test Set)",
    x     = "Specificity",
    y     = "Sensitivity"
  ) +
  annotate("text", x = 0.25, y = 0.1,
           label = paste("AUC =", round(auc(roc_obj), 4)),
           size = 5) +
  theme_minimal()
```


XGBoost achieves AUC = 0.8325 and F1 = 0.4424 on the binary test set —
a marginal AUC improvement over the logistic regression baseline (AUC = 0.8237,
F1 = 0.4435) but effectively identical F1. Recall of 0.799 means the model
correctly identifies 80% of diabetes cases, at the cost of precision of 0.306 —
meaning roughly 7 out of 10 positive predictions are false positives. This
precision-recall tradeoff is a direct consequence of `scale_pos_weight = 6.18`,
which aggressively penalizes minority class misclassification.

The near-identical performance between XGBoost and logistic regression is itself
a finding: despite XGBoost's capacity for nonlinear boundaries and feature
interactions, it adds almost no predictive lift on the binary task. This suggests
the decision boundary separating diabetic from non-diabetic patients in this
feature space is largely linear — the lifestyle and demographic features in BRFSS
do not contain the kind of complex interactions that tree-based models are
designed to exploit. The performance ceiling of approximately F1 = 0.44 and
AUC = 0.83 appears to be a property of the feature set itself, not the model
family.

### Section 2d — SHAP Analysis (Binary)

SHAP values decompose each prediction into individual feature contributions,
showing not just which features matter but in which direction and magnitude
for each observation. We compute SHAP on `binary_train` only — this is not
data snooping, we are interrogating what the model learned from training data.
The results directly inform feature engineering in Section 2e.



```{r binary-shap}

set.seed(571)

X_train_mat <- as.matrix(
  binary_train[, setdiff(names(binary_train), "Diabetes_binary")]
)


sample_idx  <- sample(nrow(X_train_mat), 10000)
X_train_sample <- X_train_mat[sample_idx, ]

shap_values <- shap.values(
  xgb_model = final_model_binary,
  X_train   = X_train_sample
)

shap_long <- shap.prep(
  xgb_model = final_model_binary,
  X_train   = X_train_sample
)

shap.plot.summary(shap_long) +
  labs(title = "SHAP Summary — Binary XGBoost (Train Set)") +
  theme_minimal()

shap_importance <- shap_values$mean_shap_score
shap_imp_df <- data.frame(
  feature    = names(shap_importance),
  mean_shap  = shap_importance
) |>
  arrange(desc(mean_shap))

ggplot(shap_imp_df, aes(x = reorder(feature, mean_shap), y = mean_shap)) +
  geom_bar(stat = "identity", fill = "#2c7bb6") +
  coord_flip() +
  labs(
    title = "Mean Absolute SHAP — Binary XGBoost",
    x     = NULL,
    y     = "Mean |SHAP Value|"
  ) +
  theme_minimal()

cat("Top 10 features by mean absolute SHAP:\n")
print(head(shap_imp_df, 10))


```
The SHAP rankings reveal the underlying structure of diabetes risk in this
dataset and deserve careful interpretation before we move to feature engineering.

**GenHlth is the dominant predictor by a wide margin** (mean |SHAP| = 0.627),
nearly 50% higher than the next feature. This is a self-reported general health
rating on a 1-5 scale — it is less a cause of diabetes than a summary of
everything the model cannot directly measure. Patients who already feel unwell
report poor general health, and that perception absorbs signal from unmeasured
clinical variables like blood glucose and HbA1c. Its dominance is a signal of
what is missing from this dataset as much as what is present.

**BMI, HighBP, and Age form a tight metabolic cluster** (SHAP 0.40–0.43),
consistent with the established pathophysiology of Type 2 diabetes. Obesity
drives insulin resistance, hypertension shares the same metabolic origins, and
aging compounds both. The near-equal SHAP values across these three features
suggest they are partially redundant — they are measuring different facets of
the same underlying metabolic risk. This is exactly where interaction terms
could add signal that raw features cannot: a patient who is obese, hypertensive,
and older is at compounding risk that a linear combination of three separate
features understates.

**HighChol rounds out the top 5** (SHAP = 0.301), again consistent with the
metabolic syndrome cluster. The fact that all five top features are
interconnected physiologically — rather than being independent risk factors —
tells us that the model is detecting a single latent risk dimension, not five
separate ones.

**The sharp drop after HighChol is meaningful.** Sex, Income, and DiffWalk
(SHAP 0.08–0.14) contribute roughly one-fifth the signal of the top features.
Everything below HighChol is secondary. Feature engineering should focus
exclusively on the top 5 — adding complexity elsewhere will not move the needle.

**The SHAP ranking almost perfectly mirrors the logistic regression odds ratios**
from the baseline notebook, where HighBP, HighChol, BMI, and GenHlth were the
dominant coefficients. This convergence across two fundamentally different model
families — one linear, one nonlinear — strongly suggests these associations are
structural properties of the data, not modeling artifacts. XGBoost is not finding
anything new because there is nothing new to find with these features alone.

The interaction terms most worth engineering are therefore: BMI × Age,
BMI × HighBP, and GenHlth × BMI — the three pairings where compounding
metabolic risk is biologically plausible and where SHAP magnitudes suggest
enough individual signal to make an interaction term worthwhile.



### Section 2e — Feature Engineering

We construct six additional features motivated by SHAP analysis and clinical
domain knowledge. SHAP identified GenHlth, BMI, HighBP, Age, and HighChol as
the five dominant features — the interaction terms below target exactly these
pairings.

- `BMI_Age` (BMI × Age): BMI and Age are SHAP ranks 2 and 4. Diabetes risk
  from obesity is age-dependent — a BMI of 30 at age 25 carries meaningfully
  different risk than at age 60. A product term captures this divergence.
- `HighBP_HighChol` (HighBP × HighChol): SHAP ranks 3 and 5. Having both
  conditions simultaneously is the core metabolic syndrome pattern and produces
  substantially higher diabetes rates than either condition alone.
- `MetabolicRisk` (HighBP + HighChol + HeartDiseaseorAttack + Stroke): An
  additive burden score from 0 to 4 capturing cumulative cardiovascular risk.
  These four features share a common physiological pathway; their sum captures
  compounding load that individual binary flags cannot.
- `AnyBadPhysicalDays` (1 if PhysHlth > 0): PhysHlth is heavily zero-inflated
  with weak signal across its full 0-30 range. The meaningful threshold is
  whether someone reports any bad physical days at all.
- `AnyBadMentalDays` (1 if MentHlth > 0): Same reasoning as above. MentHlth
  showed almost no class separation continuously; the binary any-vs-none version
  extracts what signal exists.
- `Sedentary_Obese` (1 if PhysActivity == 0 AND BMI > 28.5): Physical inactivity
  combined with above-average BMI is a well-established compound risk factor,
  clinically stronger than either condition alone. Threshold of 28.5 is the
  training set mean BMI from EDA.

Features are constructed identically on all four datasets using deterministic
transformations only — no fitting occurs, so there is no leakage risk.


```{r feature-engineering}
engineer_features <- function(df) {
  df$BMI_Age            <- df$BMI * df$Age
  df$HighBP_HighChol    <- df$HighBP * df$HighChol
  df$MetabolicRisk      <- df$HighBP + df$HighChol + df$HeartDiseaseorAttack + df$Stroke
  df$AnyBadPhysicalDays <- as.integer(df$PhysHlth > 0)
  df$AnyBadMentalDays   <- as.integer(df$MentHlth > 0)
  df$Sedentary_Obese    <- as.integer(df$PhysActivity == 0 & df$BMI > 28.5)
  df
}

binary_train_eng     <- engineer_features(binary_train)
binary_test_eng      <- engineer_features(binary_test)
multiclass_train_eng <- engineer_features(multiclass_train)
multiclass_test_eng  <- engineer_features(multiclass_test)

cat("Engineered feature set dimensions:\n")
cat("binary_train_eng:", nrow(binary_train_eng), "x", ncol(binary_train_eng), "\n")
cat("binary_test_eng: ", nrow(binary_test_eng),  "x", ncol(binary_test_eng),  "\n")
cat("New features added: BMI_Age, HighBP_HighChol, MetabolicRisk,",
    "AnyBadPhysicalDays, AnyBadMentalDays, Sedentary_Obese\n")
```


### Section 2f — Engineered Feature Model: Retrain & Evaluate

With six engineered features added, we rebuild the training matrices and
run a focused grid search over `max_depth` and `eta` only — the two
parameters most likely to shift when the feature set changes. All other
hyperparameters are held at the values selected in Section 2a. We retrain
the final model on the full engineered training set and evaluate once on
`binary_test` for direct comparison against the baseline.


```{r engineered-grid-search}
set.seed(571)

train_idx2 <- createDataPartition(binary_train_eng$Diabetes_binary, p = 0.9, list = FALSE)
tune_data2  <- binary_train_eng[ train_idx2, ]
val_data2   <- binary_train_eng[-train_idx2, ]

X_tune2 <- as.matrix(tune_data2[, setdiff(names(tune_data2), "Diabetes_binary")])
y_tune2 <- tune_data2$Diabetes_binary
X_val2  <- as.matrix(val_data2[, setdiff(names(val_data2),  "Diabetes_binary")])
y_val2  <- val_data2$Diabetes_binary

dtune2 <- xgb.DMatrix(data = X_tune2, label = y_tune2)
dval2  <- xgb.DMatrix(data = X_val2,  label = y_val2)

param_grid2 <- expand.grid(
  max_depth        = c(4, 6, 8),
  eta              = c(0.05, 0.10),
  min_child_weight = c(5, 10),
  colsample_bytree = c(0.6, 0.8)
)

cat("Total combinations:", nrow(param_grid2), "\n\n")

seeds    <- c(571, 123, 999)
results2 <- data.frame()

for (i in seq_len(nrow(param_grid2))) {
  auc_vals <- c()
  for (s in seeds) {
    set.seed(s)
    params <- list(
      objective        = "binary:logistic",
      eval_metric      = "auc",
      scale_pos_weight = spw,
      max_depth        = param_grid2$max_depth[i],
      eta              = param_grid2$eta[i],
      min_child_weight = param_grid2$min_child_weight[i],
      subsample        = 0.6,
      colsample_bytree = param_grid2$colsample_bytree[i],
      gamma            = 0.0,
      seed             = s
    )
    model <- xgb.train(
      params                = params,
      data                  = dtune2,
      nrounds               = 1000,
      evals                 = list(val = dval2),
      early_stopping_rounds = 30,
      verbose               = 0
    )
    eval_log <- attributes(model)$evaluation_log
    if (!is.null(eval_log) && "val_auc" %in% colnames(eval_log)) {
      auc_vals <- c(auc_vals, max(eval_log$val_auc))
    }
  }
  mean_auc <- mean(auc_vals, na.rm = TRUE)
  auc_sd   <- sd(auc_vals, na.rm = TRUE)
  n        <- length(auc_vals)
  ci_half  <- 1.96 * auc_sd / sqrt(n)
  results2 <- rbind(results2, data.frame(
    max_depth        = param_grid2$max_depth[i],
    eta              = param_grid2$eta[i],
    min_child_weight = param_grid2$min_child_weight[i],
    colsample_bytree = param_grid2$colsample_bytree[i],
    mean_auc         = round(mean_auc, 5),
    std_auc          = round(auc_sd, 5),
    ci_lower         = round(mean_auc - ci_half, 5),
    ci_upper         = round(mean_auc + ci_half, 5)
  ))
  cat(sprintf(
    "Combo %02d/%02d | depth=%d eta=%.2f mcw=%d col=%.1f | AUC=%.5f ± %.5f | CI [%.5f, %.5f]\n",
    i, nrow(param_grid2),
    param_grid2$max_depth[i],
    param_grid2$eta[i],
    param_grid2$min_child_weight[i],
    param_grid2$colsample_bytree[i],
    mean_auc, auc_sd,
    mean_auc - ci_half,
    mean_auc + ci_half
  ))
}

results2 <- results2[order(-results2$mean_auc), ]
cat("\nTop 5 configs:\n")
print(head(results2, 5))
cat("\nBest configuration:\n")
print(results2[1, ])
```


We select `max_depth = 4, eta = 0.10, min_child_weight = 10, colsample_bytree = 0.8`
from the top-performing cluster. All five configurations sit within overlapping
confidence intervals — no single configuration is meaningfully superior. We
prioritize the shallowest, most regularized option that matches our baseline eta.

The validation AUC of 0.8273 is virtually identical to the baseline model's 0.8275,
suggesting the six engineered features add no meaningful signal on the validation
slice. This is consistent with the SHAP finding that XGBoost already captures
feature interactions internally — manual interaction terms do not expose new
structure that the model was not already exploiting. The test evaluation in the
next section will confirm whether this holds on held-out data.



```{r engineered-final-train}
set.seed(571)

best_params_eng <- list(
  objective        = "binary:logistic",
  eval_metric      = "auc",
  scale_pos_weight = spw,
  max_depth        = 4,
  eta              = 0.10,
  min_child_weight = 10,
  subsample        = 0.6,
  colsample_bytree = 0.8,
  gamma            = 0.0,
  seed             = 571
)

dtrain_full_eng <- xgb.DMatrix(
  data  = as.matrix(binary_train_eng[, setdiff(names(binary_train_eng), "Diabetes_binary")]),
  label = binary_train_eng$Diabetes_binary
)

final_model_eng <- xgb.train(
  params  = best_params_eng,
  data    = dtrain_full_eng,
  nrounds = 136,
  verbose = 0
)

cat("Engineered model trained\n")
cat("Number of trees:", 136, "\n")
cat("Training observations:", nrow(binary_train_eng), "\n")
```


### Section 2g — Final Evaluation: Engineered Model on Test Set

The engineered model is evaluated exactly once on `binary_test_eng` —
the test set with the same six features appended identically. Results
are compared directly against the raw feature baseline to determine
whether feature engineering adds measurable lift.

```{r engineered-final-eval}
set.seed(571)

X_test_eng <- as.matrix(binary_test_eng[, setdiff(names(binary_test_eng), "Diabetes_binary")])
y_test_eng  <- binary_test_eng$Diabetes_binary

dtest_eng <- xgb.DMatrix(data = X_test_eng, label = y_test_eng)

probs_eng <- predict(final_model_eng, dtest_eng)
preds_eng <- ifelse(probs_eng > 0.5, 1, 0)

cm_eng <- confusionMatrix(
  factor(preds_eng, levels = c(0, 1)),
  factor(y_test_eng, levels = c(0, 1)),
  positive = "1"
)

print(cm_eng$table)

tp <- cm_eng$table[2, 2]
fp <- cm_eng$table[2, 1]
fn <- cm_eng$table[1, 2]

precision <- tp / (tp + fp)
recall    <- tp / (tp + fn)
f1        <- 2 * precision * recall / (precision + recall)

cat("Accuracy: ",  round(cm_eng$overall["Accuracy"], 4), "\n")
cat("Precision:",  round(precision, 4), "\n")
cat("Recall:   ",  round(recall,    4), "\n")
cat("F1:       ",  round(f1,        4), "\n")

roc_obj_eng <- roc(y_test_eng, probs_eng)
cat("ROC-AUC:  ",  round(auc(roc_obj_eng), 4), "\n")

ggroc(roc_obj_eng, color = "#2c7bb6", linewidth = 1) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray50") +
  labs(
    title = "Binary XGBoost (Engineered) — ROC Curve (Test Set)",
    x     = "Specificity",
    y     = "Sensitivity"
  ) +
  annotate("text", x = 0.25, y = 0.1,
           label = paste("AUC =", round(auc(roc_obj_eng), 4)),
           size = 5) +
  theme_minimal()
```

The engineered feature model produces virtually identical results to the raw baseline across every metric: AUC 0.8322 vs 0.8325, F1 0.4417 vs 0.4429, Recall 0.7986 vs 0.7999, Precision 0.3053 vs 0.3063. The differences are in the fourth decimal place and have no practical meaning — notably, the engineered model is marginally worse on every metric, not better.

This confirms the SHAP hypothesis: XGBoost already captures the interactions that the engineered features make explicit. BMI_Age, HighBP_HighChol, and MetabolicRisk add no new signal because the gradient boosted trees were already splitting on these feature combinations internally. Feature engineering helps models that cannot learn interactions on their own — logistic regression being the primary example. For XGBoost on this dataset, the raw features contain all the learnable signal the engineered features encode.

The performance ceiling of AUC ≈ 0.83 and F1 ≈ 0.44 is now confirmed across three models — raw LR, raw XGBoost, and engineered XGBoost. This is a feature set ceiling, not a modeling ceiling.


### Section 2h — SHAP Analysis: Engineered Feature Model

We rerun SHAP on the engineered model to determine which of the six new
features were actually used by XGBoost and whether they displaced or
complemented the original top predictors. This directly answers whether
feature engineering added interpretable structure beyond what the raw
features already provided.

```{r engineered-shap}
set.seed(571)

X_train_eng_mat <- as.matrix(
  binary_train_eng[, setdiff(names(binary_train_eng), "Diabetes_binary")]
)

sample_idx_eng    <- sample(nrow(X_train_eng_mat), 10000)
X_train_eng_sample <- X_train_eng_mat[sample_idx_eng, ]

shap_values_eng <- shap.values(
  xgb_model = final_model_eng,
  X_train   = X_train_eng_sample
)

shap_long_eng <- shap.prep(
  xgb_model = final_model_eng,
  X_train   = X_train_eng_sample
)

shap.plot.summary(shap_long_eng) +
  labs(title = "SHAP Summary — Engineered XGBoost (Train Set)") +
  theme_minimal()

shap_imp_eng_df <- data.frame(
  feature   = names(shap_values_eng$mean_shap_score),
  mean_shap = shap_values_eng$mean_shap_score
) |>
  arrange(desc(mean_shap))

ggplot(shap_imp_eng_df, aes(x = reorder(feature, mean_shap), y = mean_shap)) +
  geom_bar(stat = "identity", fill = "#2c7bb6") +
  coord_flip() +
  labs(
    title = "Mean Absolute SHAP — Engineered XGBoost",
    x     = NULL,
    y     = "Mean |SHAP Value|"
  ) +
  theme_minimal()

cat("Full feature ranking by mean absolute SHAP:\n")
print(shap_imp_eng_df)
```

The engineered SHAP rankings tell a clear and important story.

**MetabolicRisk immediately enters at rank 2** (mean |SHAP| = 0.410), displacing
BMI and HighBP from their original positions. This makes sense — MetabolicRisk
aggregates HighBP, HighChol, HeartDiseaseorAttack, and Stroke into a single
burden score, absorbing their combined signal into one feature. As a consequence,
HighBP drops from rank 3 (0.420) to rank 6 (0.175) and HighChol drops out of
the top 5 entirely (0.057). The engineered feature is doing real work — it is
just reorganizing existing signal rather than adding new signal, which explains
why AUC did not improve.

**BMI_Age enters at rank 4** (0.253), with raw BMI falling to rank 5 (0.214).
The interaction term partially replaces its parent features, again confirming
that XGBoost was already implicitly capturing this interaction — making it
explicit reshuffles importance without lifting performance.

**The bottom four engineered features are effectively dead weight.**
`HighBP_HighChol` (0.026), `AnyBadPhysicalDays` (0.004), `AnyBadMentalDays`
(0.003), and `Sedentary_Obese` (0.000) contribute near-zero SHAP. `Sedentary_Obese`
is completely ignored by the model. These features either duplicate existing
signal or contain too little independent information to be useful.

**GenHlth remains the dominant predictor** (0.657), unchanged from the baseline
model. No engineered feature comes close to displacing it, reinforcing that
self-reported general health is capturing unmeasured clinical variance that no
combination of the available lifestyle features can replicate.

The conclusion is unambiguous: feature engineering reorganized feature importance
without adding predictive value. The two features worth keeping for the multiclass
model are `MetabolicRisk` and `BMI_Age` — the rest can be dropped without any
performance cost.

## Binary XGBoost — Summary & Comparison

Across all binary models tested, performance converges to the same ceiling:

| Model | AUC | F1 | Precision | Recall | Accuracy |
|:---|:---:|:---:|:---:|:---:|:---:|
| LR (Class Weights) | 0.8237 | 0.4435 | 0.3116 | 0.7692 | 0.7310 |
| XGBoost (Raw) | 0.8325 | 0.4429 | 0.3063 | 0.7999 | 0.7196 |
| XGBoost (Engineered) | 0.8322 | 0.4417 | 0.3053 | 0.7986 | 0.7187 |

XGBoost maintains a consistent ~0.007 AUC advantage over LR, but no
model or feature engineering strategy breaks through AUC = 0.833 or
F1 = 0.445. This is a feature set ceiling, not a modeling ceiling.

The reason XGBoost does not dramatically outperform logistic regression
comes down to what SHAP revealed: the five dominant predictors —
GenHlth, BMI, HighBP, Age, HighChol — have a roughly linear relationship
to diabetes risk. XGBoost's advantage comes from capturing complex
nonlinear interactions, but when the decision boundary is already close
to linear, the tree ensemble has little extra structure to exploit.

Adding engineered features to LR would close the gap slightly —
`MetabolicRisk` and `BMI_Age` would give LR access to interactions it
cannot discover on its own. But it would not surpass XGBoost, because
XGBoost already learned those interactions from the raw features
internally. The ceiling holds regardless of model family or feature set.

**Confusion Matrix — XGBoost Baseline (Test Set)**

| | **Actual: No Diabetes** | **Actual: Diabetes** |
|---|---|---|
| **Predicted: No Diabetes** | TN = 30,846 ✅ | FN = 1,415 ❌ |
| **Predicted: Diabetes** | FP = 12,797 ⚠️ | TP = 5,655 ✅ |

The critical number is the 1,415 false negatives — diabetic patients the
model misses entirely, who leave a screening believing they are healthy.
The 12,797 false positives are a tolerable cost in a screening context:
over-flagging is correctable with follow-up testing, but a missed
diagnosis is not. Recall of 0.80 — catching 4 in 5 true cases — is the
metric that matters clinically, not the 72% accuracy figure.

Unlocking meaningful gains would require clinical biomarkers — HbA1c,
fasting glucose, insulin levels — that are absent from the BRFSS survey.
No modeling choice compensates for missing features that carry the true
discriminative signal.



### Section 3 — Multiclass XGBoost

### Overview

We extend the XGBoost pipeline to the three-class task (`Diabetes_012`:
0 = no diabetes, 1 = prediabetes, 2 = diabetes). We use the engineered
datasets throughout — `multiclass_train_eng` and `multiclass_test_eng` —
since SHAP confirmed `MetabolicRisk` and `BMI_Age` carry real signal at
no additional cost.

The key challenge here is the severe three-way class imbalance: 84% no
diabetes, 1.82% prediabetes, 14% diabetes. Unlike binary where
`scale_pos_weight` handles imbalance natively, multiclass XGBoost requires
a per-sample weight vector passed directly to `xgb.DMatrix`. Each
observation is weighted inversely proportional to its class frequency,
giving prediabetes cases approximately 46x the weight of no-diabetes cases.

The primary question is whether XGBoost's nonlinear boundaries improve
on the logistic regression multiclass baseline — particularly for
prediabetes, where LR achieved F1 = 0.056 despite a 46x class weight.


```{r multiclass-setup}
set.seed(571)

# per-sample inverse frequency weights
class_counts_mc <- table(multiclass_train_eng$Diabetes_012)
max_count_mc    <- max(class_counts_mc)

weights_mc <- as.numeric(
  max_count_mc / class_counts_mc[as.character(multiclass_train_eng$Diabetes_012)]
)


cat("Class counts:\n")
print(class_counts_mc)
cat("\nClass weights:\n")
print(round(max_count_mc / class_counts_mc, 4))

# stratified 90/10 validation slice
train_idx_mc <- createDataPartition(multiclass_train_eng$Diabetes_012, p = 0.9, list = FALSE)
tune_data_mc  <- multiclass_train_eng[ train_idx_mc, ]
val_data_mc   <- multiclass_train_eng[-train_idx_mc, ]

# weights for tune slice only
weights_tune_mc <- as.numeric(
  max_count_mc / class_counts_mc[as.character(tune_data_mc$Diabetes_012)]
)

X_tune_mc <- as.matrix(tune_data_mc[, setdiff(names(tune_data_mc), "Diabetes_012")])
y_tune_mc <- tune_data_mc$Diabetes_012
X_val_mc  <- as.matrix(val_data_mc[, setdiff(names(val_data_mc),  "Diabetes_012")])
y_val_mc  <- val_data_mc$Diabetes_012

dtune_mc <- xgb.DMatrix(data = X_tune_mc, label = y_tune_mc, weight = weights_tune_mc)
dval_mc  <- xgb.DMatrix(data = X_val_mc,  label = y_val_mc)

cat("\nTuning slice:    ", nrow(tune_data_mc), "observations\n")
cat("Validation slice:", nrow(val_data_mc),  "observations\n")
```





### Section 3b — Multiclass Grid Search

We perform a robust multi-seed grid search on the validation slice to
select the best hyperparameters for the multiclass model. The structure
is identical to the binary grid search — three seeds, early stopping on
`dval_mc`, results aggregated by mean mlogloss across seeds.

We use `mlogloss` (multiclass log loss) as the training and early
stopping metric — not AUC. This is the correct choice for multiclass
XGBoost for two reasons: first, mlogloss is smooth and differentiable,
making it stable for early stopping. Second, it aligns with how XGBoost
optimizes internally via `multi:softprob`. AUC is not used here because
it becomes noisy as an early stopping signal for multiclass problems and
does not align with the probability outputs. Instead, AUC is computed
manually in Section 3d using one-vs-rest via `pROC` — this is the staff
data scientist approach: train with mlogloss, evaluate with AUC and F1.

**Lower mlogloss is better** — it measures the penalty for incorrect or
uncertain probability predictions across all three classes simultaneously.
A perfect model would achieve mlogloss = 0. Higher values indicate the
model is either predicting the wrong class or assigning low confidence
to the correct class. We select the configuration with the lowest mean
mlogloss across seeds.

`dval_mc` is passed without weights — evaluation should reflect the true
class distribution. Weights are applied only during training.

```{r multiclass-grid-search}
set.seed(571)

param_grid_mc <- expand.grid(
  max_depth        = c(4, 6),
  eta              = c(0.05, 0.10),
  min_child_weight = c(5, 10),
  colsample_bytree = c(0.6, 0.8)
)

cat("Total combinations:", nrow(param_grid_mc), "\n\n")

seeds      <- c(571, 123, 999)
results_mc <- data.frame()

for (i in seq_len(nrow(param_grid_mc))) {
  loss_vals <- c()
  for (s in seeds) {
    set.seed(s)
    params <- list(
      objective        = "multi:softprob",
      eval_metric      = "mlogloss",
      num_class        = 3,
      max_depth        = param_grid_mc$max_depth[i],
      eta              = param_grid_mc$eta[i],
      min_child_weight = param_grid_mc$min_child_weight[i],
      subsample        = 0.6,
      colsample_bytree = param_grid_mc$colsample_bytree[i],
      gamma            = 0.0,
      seed             = s
    )
    model <- xgb.train(
      params                = params,
      data                  = dtune_mc,
      nrounds               = 1000,
      evals                 = list(val = dval_mc),
      early_stopping_rounds = 30,
      verbose               = 0
    )
    eval_log <- attributes(model)$evaluation_log
    if (!is.null(eval_log) && "val_mlogloss" %in% colnames(eval_log)) {
      loss_vals <- c(loss_vals, min(eval_log$val_mlogloss))
    }
  }
  mean_loss <- mean(loss_vals, na.rm = TRUE)
  loss_sd   <- sd(loss_vals,   na.rm = TRUE)
  n         <- length(loss_vals)
  ci_half   <- 1.96 * loss_sd / sqrt(n)
  results_mc <- rbind(results_mc, data.frame(
    max_depth        = param_grid_mc$max_depth[i],
    eta              = param_grid_mc$eta[i],
    min_child_weight = param_grid_mc$min_child_weight[i],
    colsample_bytree = param_grid_mc$colsample_bytree[i],
    mean_loss        = round(mean_loss, 5),
    std_loss         = round(loss_sd,   5),
    ci_lower         = round(mean_loss - ci_half, 5),
    ci_upper         = round(mean_loss + ci_half, 5)
  ))
  cat(sprintf(
    "Combo %02d/%02d | depth=%d eta=%.2f mcw=%d col=%.1f | mlogloss=%.5f ± %.5f | CI [%.5f, %.5f]\n",
    i, nrow(param_grid_mc),
    param_grid_mc$max_depth[i],
    param_grid_mc$eta[i],
    param_grid_mc$min_child_weight[i],
    param_grid_mc$colsample_bytree[i],
    mean_loss, loss_sd,
    mean_loss - ci_half,
    mean_loss + ci_half
  ))
}

results_mc <- results_mc[order(results_mc$mean_loss), ]
cat("\nTop 5 configs (lowest mlogloss):\n")
print(head(results_mc, 5))
cat("\nBest configuration:\n")
print(results_mc[1, ])
```


```{r multiclass-grid-search-round2}
set.seed(571)

param_grid_mc2 <- expand.grid(
  max_depth        = c(6, 8, 10),
  eta              = c(0.10, 0.15),
  min_child_weight = c(10),
  colsample_bytree = c(0.8)
)

cat("Total combinations:", nrow(param_grid_mc2), "\n\n")

seeds      <- c(571, 123, 999)
results_mc2 <- data.frame()

for (i in seq_len(nrow(param_grid_mc2))) {
  loss_vals <- c()
  for (s in seeds) {
    set.seed(s)
    params <- list(
      objective        = "multi:softprob",
      eval_metric      = "mlogloss",
      num_class        = 3,
      max_depth        = param_grid_mc2$max_depth[i],
      eta              = param_grid_mc2$eta[i],
      min_child_weight = param_grid_mc2$min_child_weight[i],
      subsample        = 0.6,
      colsample_bytree = param_grid_mc2$colsample_bytree[i],
      gamma            = 0.0,
      seed             = s
    )
    model <- xgb.train(
      params                = params,
      data                  = dtune_mc,
      nrounds               = 1000,
      evals                 = list(val = dval_mc),
      early_stopping_rounds = 30,
      verbose               = 0
    )
    eval_log <- attributes(model)$evaluation_log
    if (!is.null(eval_log) && "val_mlogloss" %in% colnames(eval_log)) {
      loss_vals <- c(loss_vals, min(eval_log$val_mlogloss))
    }
  }
  mean_loss <- mean(loss_vals, na.rm = TRUE)
  loss_sd   <- sd(loss_vals,   na.rm = TRUE)
  n         <- length(loss_vals)
  ci_half   <- 1.96 * loss_sd / sqrt(n)
  results_mc2 <- rbind(results_mc2, data.frame(
    max_depth        = param_grid_mc2$max_depth[i],
    eta              = param_grid_mc2$eta[i],
    min_child_weight = param_grid_mc2$min_child_weight[i],
    colsample_bytree = param_grid_mc2$colsample_bytree[i],
    mean_loss        = round(mean_loss, 5),
    std_loss         = round(loss_sd,   5),
    ci_lower         = round(mean_loss - ci_half, 5),
    ci_upper         = round(mean_loss + ci_half, 5)
  ))
  cat(sprintf(
    "Combo %02d/%02d | depth=%d eta=%.2f mcw=%d col=%.1f | mlogloss=%.5f ± %.5f | CI [%.5f, %.5f]\n",
    i, nrow(param_grid_mc2),
    param_grid_mc2$max_depth[i],
    param_grid_mc2$eta[i],
    param_grid_mc2$min_child_weight[i],
    param_grid_mc2$colsample_bytree[i],
    mean_loss, loss_sd,
    mean_loss - ci_half,
    mean_loss + ci_half
  ))
}

results_mc2 <- results_mc2[order(results_mc2$mean_loss), ]
cat("\nTop configs (lowest mlogloss):\n")
print(results_mc2)
cat("\nBest configuration:\n")
print(results_mc2[1, ])
```

```{r multiclass-grid-search-round3}
set.seed(571)

param_grid_mc3 <- expand.grid(
  max_depth        = c(12, 14, 16, 18, 20, 22, 24, 26, 28, 30),
  eta              = c(0.10)
)

cat("Total combinations:", nrow(param_grid_mc3), "\n\n")

seeds       <- c(571, 123)
results_mc3 <- data.frame()

for (i in seq_len(nrow(param_grid_mc3))) {
  loss_vals <- c()
  for (s in seeds) {
    set.seed(s)
    params <- list(
      objective        = "multi:softprob",
      eval_metric      = "mlogloss",
      num_class        = 3,
      max_depth        = param_grid_mc3$max_depth[i],
      eta              = 0.10,
      min_child_weight = 10,
      subsample        = 0.6,
      colsample_bytree = 0.8,
      gamma            = 0.0,
      seed             = s
    )
    model <- xgb.train(
      params                = params,
      data                  = dtune_mc,
      nrounds               = 1000,
      evals                 = list(val = dval_mc),
      early_stopping_rounds = 30,
      verbose               = 0
    )
    eval_log <- attributes(model)$evaluation_log
    if (!is.null(eval_log) && "val_mlogloss" %in% colnames(eval_log)) {
      loss_vals <- c(loss_vals, min(eval_log$val_mlogloss))
    }
  }
  mean_loss <- mean(loss_vals, na.rm = TRUE)
  loss_sd   <- sd(loss_vals,   na.rm = TRUE)
  results_mc3 <- rbind(results_mc3, data.frame(
    max_depth = param_grid_mc3$max_depth[i],
    mean_loss = round(mean_loss, 5),
    std_loss  = round(loss_sd,   5)
  ))
  cat(sprintf(
    "Combo %02d/%02d | depth=%d | mlogloss=%.5f ± %.5f\n",
    i, nrow(param_grid_mc3),
    param_grid_mc3$max_depth[i],
    mean_loss, loss_sd
  ))
}

results_mc3 <- results_mc3[order(results_mc3$mean_loss), ]
print(results_mc3)
```

The three-round grid search progressively expanded model complexity to
identify where performance gains become marginal relative to complexity cost.

**Round 1** established that `max_depth = 4` severely underfits the
multiclass problem (mlogloss ≈ 0.83–0.84) while `max_depth = 6` drops
to ≈ 0.69–0.75 — a large, non-overlapping separation. This immediately
signaled that multiclass classification requires substantially deeper
interactions than the binary task, where shallow trees dominated. Each
subsequent round was motivated by this observed trend: as long as
performance continued improving meaningfully, we kept searching.

**Round 2** confirmed the trend continued well beyond depth 6 — depth 8
(0.586) and depth 10 (0.556) both improved with tight, stable CIs.
`eta = 0.15` was tested and consistently underperformed `eta = 0.10` at
every depth and was eliminated. No plateau was visible, justifying Round 3.

**Round 3** stress-tested depths 12–30, finding genuine saturation
between depth 24–28. The curve flattens sharply after depth 20 (0.505)
— depth 22 gains 0.004, depth 24 gains 0.002, depth 26 gains 0.001,
depth 28 gains 0.002. These differences are within noise.

**Why this problem requires deeper trees than binary.** Three class
boundaries are inherently more complex than one. The 1.82% prediabetes
class introduces severe asymmetry that requires fine-grained splits to
detect. And the interactions between health indicators — BMI, Age, HighBP,
GenHlth — are not linearly separable across three classes the way they
are across two.

**Selected configuration: `max_depth = 20, eta = 0.10,
min_child_weight = 10, subsample = 0.6, colsample_bytree = 0.8`**

We select depth 20 over the numerically optimal depth 28 deliberately.
The mlogloss difference is 0.004 — well within noise. Against this
negligible gain, depth 28 introduces: extreme interpretability costs
(SHAP becomes unreliable at this depth), elevated overfitting risk on a
200k dataset, substantially slower training, and a model that is not
defensible in an industry context without a meaningful performance
justification. Depth 20 sits at the Pareto-optimal point — near-optimal
performance with a model that remains interpretable, stable, and
production-worthy. This is the correct staff-level tradeoff.


### Section 3c — Final Multiclass Model Training on Full Training Set

With `max_depth = 20` selected as the Pareto-optimal configuration, we
retrain the final model on the complete `multiclass_train_eng` dataset.
The per-sample weight vector is recomputed on the full training set to
ensure all 202,943 observations are correctly weighted. We use the
round-finder approach to determine optimal nrounds via early stopping on
the validation slice, then hardcode and retrain clean on the full data
with no held-out slice — consistent with the binary pipeline.

```{r multiclass-final-train}
set.seed(571)

best_params_mc <- list(
  objective        = "multi:softprob",
  eval_metric      = "mlogloss",
  num_class        = 3,
  max_depth        = 20,
  eta              = 0.10,
  min_child_weight = 10,
  subsample        = 0.6,
  colsample_bytree = 0.8,
  gamma            = 0.0,
  seed             = 571
)

weights_full_mc <- as.numeric(
  max_count_mc / class_counts_mc[as.character(multiclass_train_eng$Diabetes_012)]
)

dtrain_full_mc <- xgb.DMatrix(
  data   = as.matrix(multiclass_train_eng[, setdiff(names(multiclass_train_eng), "Diabetes_012")]),
  label  = multiclass_train_eng$Diabetes_012,
  weight = weights_full_mc
)

final_model_mc <- xgb.train(
  params  = best_params_mc,
  data    = dtrain_full_mc,
  nrounds = 106,
  verbose = 0
)

cat("Final multiclass model trained\n")
cat("Number of trees:", 106, "\n")
cat("Training observations:", nrow(multiclass_train_eng), "\n")
```


### Section 3d — Final Evaluation on Multiclass Test Set

The final multiclass model is evaluated exactly once on `multiclass_test_eng`
— data never touched during training, validation slicing, or hyperparameter
tuning. We report a full confusion matrix heatmap, per-class precision,
recall, and F1, macro and weighted F1, and one-vs-rest ROC-AUC per class.

The primary question is whether XGBoost's deeper nonlinear boundaries
improve on the logistic regression multiclass baseline — particularly for
prediabetes (X1), where LR achieved F1 = 0.056 despite a 46x class weight.
Aggregate metrics like accuracy and weighted F1 will be dominated by the
majority class — per-class metrics, especially X1, are what matter here.

Predictions are generated using `multi:softprob` which outputs a probability
vector of length 3 per observation. The predicted class is the argmax of
these probabilities. ROC-AUC is computed manually using `pROC` one-vs-rest
for each class and averaged.

```{r multiclass-final-eval}
set.seed(571)

X_test_mc <- as.matrix(
  multiclass_test_eng[, setdiff(names(multiclass_test_eng), "Diabetes_012")]
)
y_test_mc <- multiclass_test_eng$Diabetes_012

dtest_mc <- xgb.DMatrix(data = X_test_mc, label = y_test_mc)

probs_mc  <- predict(final_model_mc, dtest_mc)
probs_mat <- matrix(probs_mc, ncol = 3)
colnames(probs_mat) <- c("X0", "X1", "X2")

preds_mc <- max.col(probs_mat) - 1

cm_mc <- confusionMatrix(
  factor(preds_mc, levels = c(0, 1, 2)),
  factor(y_test_mc, levels = c(0, 1, 2))
)

print(cm_mc$table)

cat("\nPer-class metrics:\n")
per_class <- cm_mc$byClass[, c("Precision", "Recall", "F1")]
print(round(per_class, 4))

macro_f1    <- mean(cm_mc$byClass[, "F1"], na.rm = TRUE)
counts_test <- table(y_test_mc)
weighted_f1 <- sum(
  cm_mc$byClass[, "F1"] * as.numeric(counts_test) / sum(counts_test),
  na.rm = TRUE
)

cat("\nMacro F1:   ", round(macro_f1,    4), "\n")
cat("Weighted F1:", round(weighted_f1, 4), "\n")
cat("Accuracy:   ", round(cm_mc$overall["Accuracy"], 4), "\n")

roc_x0 <- roc(ifelse(y_test_mc == 0, 1, 0), probs_mat[, "X0"], quiet = TRUE)
roc_x1 <- roc(ifelse(y_test_mc == 1, 1, 0), probs_mat[, "X1"], quiet = TRUE)
roc_x2 <- roc(ifelse(y_test_mc == 2, 1, 0), probs_mat[, "X2"], quiet = TRUE)

cat("\nROC-AUC per class (one-vs-rest):\n")
cat("X0 (No Diabetes):", round(auc(roc_x0), 4), "\n")
cat("X1 (Prediabetes):", round(auc(roc_x1), 4), "\n")
cat("X2 (Diabetes):   ", round(auc(roc_x2), 4), "\n")
cat("Average AUC:     ", round(mean(c(auc(roc_x0), auc(roc_x1), auc(roc_x2))), 4), "\n")

cm_df <- as.data.frame(cm_mc$table)
colnames(cm_df) <- c("Predicted", "Actual", "Count")

cm_df <- cm_df |>
  group_by(Actual) |>
  mutate(
    RowPct  = round(Count / sum(Count) * 100, 1),
    Actual  = recode(Actual,  "0" = "No Diabetes", "1" = "Prediabetes", "2" = "Diabetes"),
    Predicted = recode(Predicted, "0" = "No Diabetes", "1" = "Prediabetes", "2" = "Diabetes")
  ) |>
  ungroup()

ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(Count, "\n(", RowPct, "%)")), size = 4) +
  scale_fill_gradient(low = "#deebf7", high = "#2c7bb6") +
  labs(
    title = "Confusion Matrix — Multiclass XGBoost (Test Set)",
    x     = "Predicted Class",
    y     = "Actual Class"
  ) +
  theme_minimal()
```

### Section 3d Analysis — Multiclass Test Set Results

XGBoost achieves meaningful improvements over logistic regression on the
majority classes but fails to improve prediabetes detection — the hardest
and most clinically important boundary in this dataset.

| Metric       | LR    | XGBoost |
| ------------ | ----- | ------- |
| X0 Precision | 0.947 | 0.901   |
| X0 Recall    | 0.670 | 0.851   |
| X0 F1        | 0.785 | 0.875   |
| X0 AUC       | 0.813 | 0.789   |
| X1 Precision | 0.031 | 0.036   |
| X1 Recall    | 0.295 | 0.016   |
| X1 F1        | 0.056 | 0.022   |
| X1 AUC       | 0.680 | 0.638   |
| X2 Precision | 0.353 | 0.362   |
| X2 Recall    | 0.577 | 0.511   |
| X2 F1        | 0.438 | 0.424   |
| X2 AUC       | 0.818 | 0.796   |
| Macro F1     | 0.426 | 0.440   |
| Weighted F1  | 0.723 | 0.796   |
| Avg AUC      | 0.770 | 0.741   |
| Accuracy     | 0.650 | 0.788   |


**No Diabetes (X0):** XGBoost substantially improves recall (0.851 vs
0.670) at the cost of slightly lower precision (0.901 vs 0.947),
producing a higher F1 of 0.875 vs 0.785. XGBoost correctly clears 85%
of healthy patients vs LR's 67% — a meaningful clinical improvement.

**Diabetes (X2):** Nearly identical performance across both models.
XGBoost gains slightly on precision (0.362 vs 0.353) but loses on
recall (0.511 vs 0.577), producing effectively the same F1 (0.424 vs
0.438). Neither model meaningfully separates diabetes from no-diabetes
beyond what the raw feature signal allows.

**Prediabetes (X1):** XGBoost is worse on every metric. Recall
collapses from 0.295 to 0.016 — LR identifies 30% of prediabetes cases
while XGBoost identifies only 1.5%. Precision is nearly identical and
near-zero for both. AUC drops from 0.680 to 0.638. This is the critical
finding: despite deeper trees, engineered features, and 46x sample
weights, XGBoost is less capable of detecting prediabetes than logistic
regression. The reason is optimization objective — mlogloss on a heavily
imbalanced three-class problem gets dominated by the majority classes,
effectively teaching the model to ignore prediabetes. LR with explicit
class weights directly penalizes prediabetes misclassification in a way
that mlogloss does not.

**The convergence of two model families on the same failure point is
itself the finding.** Both models achieve X1 F1 ≈ 0.02–0.06 despite
fundamentally different architectures. This is not a modeling failure —
it is a feature set limitation. Clinical biomarkers — HbA1c, fasting
glucose, oral glucose tolerance — are the actual diagnostic criteria
separating prediabetes from diabetes and no-diabetes. Their absence from
BRFSS makes this boundary effectively unlearnable from lifestyle
features alone, regardless of model complexity.

**Confusion Matrix — XGBoost Multiclass (Test Set)**

| | **Actual: No Diabetes** | **Actual: Prediabetes** | **Actual: Diabetes** |
|---|---|---|---|
| **Predicted: No Diabetes** | TN = 36,258 ✅ | FN = 597 ❌ | FN = 3,351 ❌ |
| **Predicted: Prediabetes** | FP = 358 ⚠️ | TP = 14 ✅ | FP = 66 ⚠️ |
| **Predicted: Diabetes** | FP = 6,112 ⚠️ | FN = 327 ❌ | TP = 3,654 ✅ |

Of 938 true prediabetes cases, only 14 are correctly identified —
a 1.5% detection rate. The remaining 924 split between no-diabetes
(597) and diabetes (327), confirming the model has essentially no
ability to distinguish prediabetes from adjacent classes. The 6,112
healthy patients incorrectly flagged as diabetic represent the largest
source of error, driven by the model's aggressive weighting toward
the diabetes class.


### Section 3e — SHAP Analysis: Multiclass Model

We compute SHAP values for the multiclass model to understand what
XGBoost is learning about each class — particularly prediabetes, where
the model failed almost completely. Unlike binary SHAP which produces
one value per feature per observation, multiclass SHAP produces three
values per feature per observation — one for each class. This lets us
ask: which features push predictions toward each class, and does
XGBoost find any signal that separates prediabetes from its neighbors?
We compute on a 1k sample from `multiclass_train_eng` for speed.

```{r multiclass-shap}
set.seed(571)

X_train_mc_mat <- as.matrix(
  multiclass_train_eng[, setdiff(names(multiclass_train_eng), "Diabetes_012")]
)

sample_idx_mc     <- sample(nrow(X_train_mc_mat), 1000)
X_train_mc_sample <- X_train_mc_mat[sample_idx_mc, ]

dtrain_mc_sample <- xgb.DMatrix(data = X_train_mc_sample)

shap_raw <- predict(final_model_mc, dtrain_mc_sample, predcontrib = TRUE)

cat("SHAP array dimensions:", dim(shap_raw), "\n")

feature_names <- colnames(X_train_mc_sample)
n_features    <- length(feature_names)

shap_class0 <- shap_raw[, 1, 1:n_features]
shap_class1 <- shap_raw[, 2, 1:n_features]
shap_class2 <- shap_raw[, 3, 1:n_features]

colnames(shap_class0) <- feature_names
colnames(shap_class1) <- feature_names
colnames(shap_class2) <- feature_names

imp_class0 <- sort(colMeans(abs(shap_class0)), decreasing = TRUE)
imp_class1 <- sort(colMeans(abs(shap_class1)), decreasing = TRUE)
imp_class2 <- sort(colMeans(abs(shap_class2)), decreasing = TRUE)

cat("\nTop 10 features — Class 0 (No Diabetes):\n")
print(round(head(imp_class0, 10), 4))

cat("\nTop 10 features — Class 1 (Prediabetes):\n")
print(round(head(imp_class1, 10), 4))

cat("\nTop 10 features — Class 2 (Diabetes):\n")
print(round(head(imp_class2, 10), 4))
```
### Section 3e Analysis — Multiclass SHAP

The per-class SHAP rankings show how XGBoost distributes predictive signal across the three outcome classes and help explain the asymmetric performance observed in evaluation metrics.

**Class 0 (No Diabetes)** is primarily driven by `BMI_Age` (0.426), `GenHlth` (0.354), and `BMI` (0.298), consistent with the dominant metabolic structure seen in the binary model. These variables form a stable and highly predictive feature cluster, indicating that the model has a well-defined representation of the non-diabetic population. `MetabolicRisk` (0.256) and `Income` (0.200) also contribute meaningfully, suggesting that both physiological and socioeconomic factors are used in separating this class.

**Class 2 (Diabetes)** shows a very similar structure, with `GenHlth` (0.355) and `BMI_Age` (0.263) as the strongest contributors, followed by `BMI` and `HighBP`. The relatively higher importance of `HighBP` compared to Class 0 is consistent with established clinical relationships between hypertension and diagnosed diabetes. Overall, the Class 2 feature pattern closely mirrors the binary classification results, indicating that diabetes is the most clearly separable class in the dataset.

**Class 1 (Prediabetes) tells the most important story.** The top features are `BMI_Age` (0.667) and `BMI` (0.521), both substantially larger than in the other classes. However, this does not indicate the presence of strong prediabetes-specific signal. Instead, it reflects the absence of separable structure: when a class lacks distinct feature patterns, tree-based models tend to rely more heavily on globally predictive variables that are useful across all classes.

This interpretation is supported by the lack of any feature that is uniquely elevated for Class 1 relative to Classes 0 and 2. `Income` (0.332) and `Education` (0.209) are slightly higher for Class 1 than Class 2, but the differences are small and not sufficient to indicate a robust or clinically meaningful separation mechanism.

**Comparison with binary SHAP.** The dominant binary features — `GenHlth`, `BMI`, `HighBP`, `Age`, and `HighChol` — are all present across the multiclass SHAP results, but their influence is redistributed across classes. `HighBP` is more influential for Class 2 than Class 1, consistent with hypertension being more strongly associated with diagnosed diabetes. `GenHlth` remains important for Classes 0 and 2 but is less discriminative for Class 1.

Overall, the SHAP analysis reinforces the evaluation results: there is no clearly learnable boundary for prediabetes within this feature space. The model does not identify a distinct feature signature for Class 1, instead interpolating between the patterns learned for Classes 0 and 2.

---

```{r}

# --- XGBoost Binary Task Comparison ---
xgb_binary_df <- data.frame(
  Model      = c("XGBoost", "XGBoost"),
  Features   = c("Raw", "Engineered"),
  Val_AUC    = c(0.8275, 0.8273),
  Trees      = c(167, 136),
  Test_Acc   = c("0.7196", "0.7187"),
  Test_Prec  = c("0.3063", "0.3053"),
  Test_Rec   = c("0.7999", "0.7986"),
  Test_F1    = c("0.4429", "0.4417"),
  Test_AUC   = c("0.8325", "0.8322")
)

xgb_binary_df |>
  kbl(
    col.names = c("Model", "Feature Set",
                  "Val AUC", "Trees",
                  "Test Accuracy", "Test Precision",
                  "Test Recall", "Test F1", "Test ROC-AUC"),
    caption   = "Table X. XGBoost Binary Task — Raw vs Engineered Features.
                 Selected model highlighted in green.",
    align     = "llcccccccc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE
  ) |>
  row_spec(2, background = "#d4edda") |>   # selected model (engineered)
  column_spec(3, bold = TRUE) |>           # Val AUC
  column_spec(8, bold = TRUE)              # Test F1

```


```{r}

# --- XGBoost Multiclass Overall ---
xgb_mc_df <- data.frame(
  Model    = c("XGBoost (Engineered)"),
  Depth    = c(20),
  Trees    = c(106),
  Test_Acc = c("0.7880"),
  MacroF1  = c("0.4400"),
  WtedF1   = c("0.7960"),
  AvgAUC   = c("0.7410")
)

xgb_mc_df |>
  kbl(
    col.names = c("Model", "Max Depth", "Trees",
                  "Test Accuracy", "Macro F1",
                  "Weighted F1", "Avg AUC"),
    caption   = "Table X. XGBoost Multiclass Task — Overall Performance.",
    align     = "llccccc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) |>
  row_spec(1, background = "#d4edda") |>
  column_spec(5, bold = TRUE) |>           # Macro F1
  column_spec(7, bold = TRUE)              # Avg AUC


# --- XGBoost Multiclass Per-Class Performance ---
xgb_mc_class_df <- data.frame(
  Class     = c("X0 — No Diabetes", "X1 — Prediabetes", "X2 — Diabetes"),
  Precision = c(0.9010, 0.0360, 0.3620),
  Recall    = c(0.8510, 0.0160, 0.5110),
  F1        = c(0.8750, 0.0220, 0.4240),
  ROC_AUC   = c(0.7890, 0.6380, 0.7960)
)

xgb_mc_class_df |>
  kbl(
    col.names = c("Class", "Precision", "Recall", "F1", "ROC-AUC"),
    caption   = "Table X. XGBoost Multiclass — Per-Class Test Performance.",
    digits    = 4,
    align     = "lcccc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) |>
  row_spec(2, background = "#ffe8e8") |>   # X1 — highlight failure
  column_spec(4, bold = TRUE) |>           # F1
  column_spec(5, bold = TRUE)              # ROC-AUC

```


```{r}

library(kableExtra)
library(dplyr)

binary_block <- data.frame(
  Binary_Metric = c("Accuracy", "Precision", "Recall", "F1", "ROC-AUC",
                    "", "", "", "", "", "", "", "", "", "", ""),
  Binary_LR     = c("0.7310", "0.3116", "0.7692", "0.4435", "0.8237",
                    "", "", "", "", "", "", "", "", "", "", ""),
  Binary_XGB    = c("0.7187", "0.3053", "0.7986", "0.4417", "0.8322",
                    "", "", "", "", "", "", "", "", "", "", "")
)

mc_block <- data.frame(
  MC_Metric = c(
    "Accuracy", "Macro F1", "Weighted F1", "Avg AUC",
    "X0 Precision", "X0 Recall", "X0 F1", "X0 AUC",
    "X1 Precision", "X1 Recall", "X1 F1", "X1 AUC",
    "X2 Precision", "X2 Recall", "X2 F1", "X2 AUC"
  ),
  MC_LR = c(
    "0.650", "0.426", "0.723", "0.770",
    "0.947", "0.670", "0.785", "0.813",
    "0.031", "0.295", "0.056", "0.680",
    "0.353", "0.577", "0.438", "0.818"
  ),
  MC_XGB = c(
    "0.788", "0.440", "0.796", "0.741",
    "0.901", "0.851", "0.875", "0.789",
    "0.036", "0.016", "0.022", "0.638",
    "0.362", "0.511", "0.424", "0.796"
  )
)

combined <- cbind(binary_block, mc_block)

combined |>
  kbl(
    col.names = c(
      "Metric", "Logistic Regression", "XGBoost",
      "Metric", "Logistic Regression", "XGBoost"
    ),
    caption = "Model Comparison — Binary (left) and Multiclass (right). Test set results. XGBoost = Engineered feature set.",
    align   = "lcclcc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE
  ) |>
  row_spec(0, bold = TRUE) |>
  add_header_above(c(
    "Binary Classification" = 3,
    "Multiclass Classification" = 3
  ), bold = TRUE, background = "#2c3e50", color = "white")

```



---

## Key Findings & Interpretation

**1. XGBoost improves overall classification performance compared to logistic regression, but gains are concentrated in the majority class.**
In the multiclass task, the improvement in accuracy (0.788 vs 0.650)
and macro F1 (0.440 vs 0.426) over the logistic regression baseline is
driven primarily by better separation of Class 0 (F1 0.875 vs 0.785),
while minority class performance remains largely unchanged or regresses
— prediabetes recall collapses from 0.295 to 0.016. In the binary task,
XGBoost holds a marginal AUC advantage (0.8325 vs 0.8237) but F1 is
effectively identical (0.4429 vs 0.4435).

**2. The decision boundary is largely linear in structure.**
SHAP rankings from the binary XGBoost model closely mirror the structure
of logistic regression coefficients, with `GenHlth`, `BMI`, `HighBP`,
and `Age` dominating both models. In the engineered binary model,
`HighChol` is largely absorbed into `MetabolicRisk` and drops out of
the top predictors entirely (SHAP = 0.057), but this reorganizes
existing signal rather than revealing new nonlinear structure.

**3. Prediabetes is not well separable in this dataset.**
In the multiclass task, both models achieve consistently low performance
on Class 1 (F1 ≈ 0.02–0.06), indicating that prediabetes does not form
a distinct, learnable region in the current feature space. The binary
task sidesteps this boundary entirely by collapsing prediabetes into the
positive class, which is why binary AUC (0.83) substantially exceeds
the multiclass average AUC (0.74).

**4. Feature engineering does not materially change separability.**
In the binary task, constructed interaction features (e.g., `BMI_Age`,
`MetabolicRisk`) improve interpretability and feature ranking structure
but do not significantly change predictive ceilings — the engineered
binary model is marginally worse than raw on every metric. The
multiclass model was trained on the engineered feature set, where the
same ceiling holds.

**5. Dataset limitations dominate model performance.**
The BRFSS dataset is based on self-reported survey data, introducing
measurement error in both predictors and labels. This likely attenuates
separability across all models and tasks, imposing a hard upper bound
on achievable performance regardless of model family or feature set.

**Final takeaway:** model choice has limited impact relative to data
limitations; the primary constraint is the absence of clinically precise
biomarkers required to distinguish prediabetes from adjacent classes.





  
  
